---
title: |
 | Cosine similarity and keyword selection
 | in multidimensional semantic analysis
# author: "Alexander T. Kindel^[Assistant Professor of Sociology, médialab, Sciences Po, Paris, France. Contact: alexander.kindel@sciencespo.fr. Replication materials are available at https://github.com/atkindel/cosine_similarity. I am grateful to audiences at Princeton University, Sciences Po, and Ecole Polytechnique for helpful feedback and conversation on prior versions of this paper. I also wish to thank Anna Berg, Dahyun Choi, Tim Hansen, Jeff Lockhart, Armin Pournaki, Benjamin Rohr, Nicolas Torres-Echevarry, and Fred Wherry for critical commentary on prior drafts; Ning Zhou and Nathan West for introducing me to an aleatoric representation of multidimensional word association problems; and Brandon Stewart for extensive guidance over the lifespan of this project. Portions of this project were supported by a French government grant managed by the Agence Nationale de la Recherche under the France 2030 program, reference ANR-22-EXES-0014.]"
author: "Alexander T. Kindel^[Assistant Professor of Sociology, médialab, Sciences Po, Paris, France. Contact: alexander.kindel@sciencespo.fr.]"
date: "10 February 2025"
abstract: |
 | The association between two latent concepts is a common quantity of interest in studies of text corpora in the social sciences. Many proposed measures for this quantity construct  keyword lists representing the target concepts, obtain a vector-valued representation for each keyword, and compute a mean cosine similarity (MCS) between pairs of word vectors across the two sets. MCS-based measures depend on the level of association within each set of keyword vectors in addition to the targeted between-set association unless it is assumed that the keyword selection mechanism induces isotropic sets in expectation. The plausibility of this assumption is evaluated in a set of exemplary multidimensional semantic association problems constructed from keyword lists used in prior applications of word embeddings in the social sciences. The validity of MCS-based quantities is limited by the fact that keyword selection induces highly anisotropic sets of word vectors; in particular, 75% of keyword list pairs induce a set of between-set cosine similarities that would be geometrically impossible if the inducing keyword vectors were isotropic. Researchers should instead use an association measure that is invariant to anisotropy by construction, such as the cosine similarity between the orthogonal projection matrices induced by each set of word vectors.
 |
bibliography: Kindel_GCD_Spr24.bib
# csl: cup-journal.cls
toc: FALSE
indent: TRUE
# documentclass: cup-journal
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: TRUE
    toc: FALSE
    dev: png
    number_sections: TRUE
header-includes:
  - \usepackage{rotating}
  - \usepackage{graphicx}
  - \usepackage{setspace}
  - \usepackage{enumitem}
  - \setlist{listparindent=\parindent, parsep=0pt}
  - \doublespacing
#   - \usepackage{biblatex}
#   - \usepackage{indentfirst}
  - \setlength\parindent{24pt}
#   - \setlength{\parskip}{0.5mm}
#   - \usepackage{epigraph}
#   - \usepackage{cancel}
#   - \usepackage{bbm}
  - \usepackage{caption}
  - \captionsetup[figure]{labelfont=bf}
#   - \setlength\epigraphwidth{0.9\textwidth}
#   - \setlength\epigraphrule{0pt}
#   - \renewcommand{\textflush}{flushepinormal}
#   - \renewcommand{\epigraphflush}{center}
#   - \setlist{listparindent=\parindent, parsep=0pt}
#   - \doublespacing
#   - \usepackage{indentfirst}
#   - \setlength\parindent{24pt}
#   - \setlength{\parskip}{0.5mm}
  - \usepackage[bottom]{footmisc}
  - \renewcommand{\footnotelayout}{\setstretch{1.05}}
#   - \usepackage{dcolumn}
#   - \usepackage{hhline}
  - \usepackage{longtable}
  - \usepackage{float}
  - \setcounter{section}{1}
#
#
# - \\footnote{Assistant Professor of Sociology, médialab, Sciences Po, Paris, France. Contact: alexander.kindel@sciencespo.fr. I am grateful to audiences at Princeton University, Sciences Po, and Ecole Polytechnique for helpful feedback and conversation on prior versions of this paper. I also wish to thank A. Berg, D. Choi, T. Hansen, J. Lockhart, A. Pournaki, B. Rohr,  N. Torres-Echevarry, and F. Wherry for critical feedback on prior drafts; N. Zhou and N. West for introducing me to an aleatoric representation of multidimensional word association problems; and B. Stewart for extensive intellectual guidance over the lifespan of this project. Portions of this project were supported by a French government grant managed by the Agence Nationale de la Recherche under the France 2030 program, reference ANR-22-EXES-0014.}
# - KEYWORDS: cosine similarity, word embeddings, matrix correlation, correlation matrix, canonical correlation, anisotropy
---

\pagebreak

Many research goals in the social sciences require measures of association between latent concepts. Varieties of this problem arise in the study of sentiments, biases, stereotypes, schemas, ideologies, and other symbolic relations that are widely thought to shape how people communicate with one-another. Distributional semantic analysis is a popular way to ground claims about these relations in empirical observations of language use. In this approach, the researcher samples a corpus of relevant text; specifies a measure of co-occurrence between pairs of words in the sample vocabulary; employs numerical algorithms to represent each word's co-occurrence pattern in a latent vector space of reduced dimension (e.g., word embeddings); and chooses a measure of association between pairs of vectors in this space (e.g., cosine similarity). 

Defining a cosine similarity as the quantity of interest requires obtaining a vector-valued representation of the lexical variation of interest. In the simplest case, each target concept is identified unambiguously with a single word, and the researcher need only compute the cosine similarity between the corresponding word vectors. Unfortunately, the choice of representation is not so simple in practice: few concepts of theoretical interest in the social sciences can be represented adequately and unambiguously by a single word. In such cases the researcher must further decide which words to use for each concept, as well as how to aggregate the corresponding set of word vectors into a representation for each concept. A common strategy is to select a list of keywords that are related in the sense meant by the target concept; choose some vector-valued aggregation of the corresponding word vectors to represent each target concept; and interpret the cosine similarity between these aggregated vectors as a measure of conceptual association.

Assuming some choice of keywords is unavoidable, the validity of this measurement strategy depends principally on how well the choice of aggregation isolates the variation of interest. In particular, comparing two sets of word vectors creates a linear system in which the position of each word vector varies both *within-set* and *between-set*, but not all of this variation is necessarily of interest. When the measurement goal is the level of association *between* two concepts, it is important to ensure that the representation for each concept is invariant to the level of within-set association, because such measures should not depend on how tightly interrelated our particular choice of keywords for a given concept happens to be. This invariance principle is crucial in this setting because any admissible choice of keyword list will contain words that are more interrelated than we would expect if it were drawn uniformly at random from the sample vocabulary.^[Keyword selection defies simple characterization, and systematically characterizing the selection mechanism is beyond the scope of this paper. The findings in this paper rely only on the observation that this mechanism is highly non-random, so the expected level of within-set association for any reasonable keyword list is greater than zero. We return to the problem of keyword selection at the end of the paper.] This implies that the expected level of within-set association is nonzero. Moreover, in settings where the choice of keyword list is not uniquely determined by the choice of concept, the researcher is free to choose an element of the set of admissible keyword lists with relatively higher or lower within-set association. Choices of aggregation that are not invariant to within-set association allow this form of researcher discretion to influence the measure.

Sensitivity to within-set association limits the validity of many proposed choices of aggregation under keyword selection. In particular, a significant body of prior work recommends targeting either the cosine similarity between the centroids of each set of word vectors or the arithmetic *mean cosine similarity* (MCS) over all possible pairs of words between the two lists. The quantity is then interpreted as a measure of the association between the underlying concepts. Since all quantities constructed in this way are equivalent to MCS up to a choice of weights, this family of methods is called "MCS-based" throughout the paper [for a recent review see @stoltz2023tool; see also @kozlowski2019geometry; @nelson2021leveraging].

Although MCS-based methods are appealingly simple, their applicability as measures of conceptual association is limited by their sensitivity to the level of association within the two sets of word vectors induced by keyword selection. At a high-level, this sensitivity arises from the fact that the distribution of the set of between-set cosine similarities is constrained by the corresponding sets of within-set cosine similarities, which are fixed by keyword selection in advance of the target comparison. Mechanically, keyword selection induces comparisons between highly anisotropic subsets of the input word vector space; that is, the component vectors are far from orthogonal and are not equivalent in scale. Repeatedly resampling the keyword lists will not induce isotropic subsets of vectors on average, so the anisotropic constraint on the distribution of the between-set cosine similarities does not vanish in expectation. This means that MCS-based association measures can be distorted by choosing more or less interrelated keywords.

To avoid measurement distortion due to keyword selection, it is necessary to choose an aggregated representation for each set of word vectors that is invariant to anisotropy by construction. A particularly good choice of representation is the linear subspace spanned by each set of word vectors. Each subspace is uniquely identified with an orthogonal projection matrix, and the cosine similarity between the equivalent vectors in $\mathbb{R}^{d \times d}$ --- often called the generalized coefficient of determination (GCD) --- is a measure of multivariate association with the desired invariance property.

The goals of this paper are (1) to explain how sensitivity to within-set association arises in MCS-based methods under keyword selection and (2) to recommend an alternative choice of representation (i.e., orthogonal projection matrices) with the desired invariance property. We proceed as follows. Section 2 reviews the standard model for distributional semantic analysis and characterizes the behavior of cosine similarity in settings involving more than two word vectors. Section 3 compares cosine similarity-based methods for multidimensional associations. In this section we characterize how within-set association affects MCS-based methods, and we explain how orthogonal projections factor out this variation. Section 4 employs keyword lists drawn from prior work on word embeddings in the social sciences to compare the performance of MCS and GCD. Section 5 concludes by emphasizing the need for more extensive construct validation of text-based measurement models in the social sciences.


# Cosine similarity in distributional semantic models

Distributional semantic models provide an infrastructure for linking research questions about linguistically-mediated social interaction to empirically observable patterns of word use in text. In distributional semantic analysis, we define a *semantic space* in which the distribution of the use of each word in a large text corpus is assigned a real-valued vector, and various notions of association between two words can be measured by comparing their respective vectors in some way. The fundamental ingredients in these models are two functions $\{X, d\}$. The co-occurrence function $X$ represents the distribution of uses of each word as a real-valued multidimensional vector, and the association function $d$ defines a way of comparing two word vectors. The properties of the association function are of central interest in this section; for completeness, we also briefly review the properties of the co-occurrence function.

## Word co-occurrence function

The **word co-occurrence function** $X: \mathcal{V} \times \mathcal{V} \rightarrow \mathbb{R}$ is a bivariate function that assigns a real number to every pair of words in the sample vocabulary $\mathcal{V}$, the set of words in the analytic corpus. Typically $X$ is chosen as a function of co-occurrence frequency, for example probability-based or information-based quantities. It can be represented as a matrix with $n \times n$ entries where $n = |\mathcal{V}|$ is the cardinality of the sample vocabulary. Each column vector in $x \in X$ represents the distribution of word co-occurrences of a word $v \in \mathcal{V}$ (i.e. $x$ is a "word vector"). The working definition of "co-occurrence frequency" depends on choosing a working definition of "context"; the most commonly used co-occurrence functions define context as a window of words around each word position in the text. Other pre-processing decisions that convert raw text into an analytic corpus (e.g. including bigrams, removing stopwords) affect the function through the context mechanism, for example by moving words closer or further apart.^[Problems relating to defining context, sampling text corpora, or preprocessing the sample vocabulary are out of scope for the purposes of this paper, and readers may assume these choices are provisionally fixed. For a more developed discussion of the issues that arise in text preprocessing, see @denny2018text.]

In applications involving large text corpora, the size of the sample vocabulary makes it impractical to explicitly represent a useful co-occurrence function. To avoid incurring the cost of working with $X$ directly, researchers instead compute a low rank approximation of the form $X = C^TW$, where $W, C$ are $d \times n$ matrices and $d \ll n$. Word embedding algorithms are a popular way of constructing such an approximation. Different word embedding algorithms imply different choices of co-occurence function, so word vectors estimated using different algorithms tend to exhibit somewhat different semantic properties [@levy2014neural; @rodriguez2022word]. However, it is important to note that the measurement problem discussed in this paper applies to *any* multidimensional comparison of two sets of vectors labeled by two sets of related words, and it does not depend on a specific choice of co-occurrence function. In general we would not expect any coherent distributional semantic model to learn an orthogonal set of vectors to represent a priori related words.^[For clarity of exposition, the paper focuses on the word-level association regime, but vector space models for sentences, documents, and other forms of text are also common. The geometric constraints discussed in this paper apply to $2 \times 2$ block cosine similarity matrices regardless of which lexical unit of analysis is represented in the vector space under consideration. The crucial ingredient is the selection of two sets of vectors such that each set represents a latent concept and the variation of interest is the association between the concepts.]

## Word association function

The **word association function** $d: X \times X \rightarrow \mathbb{R}$ is a symmetric bivariate function comparing two unidimensional word co-occurrence distributions. $d$ is superficially similar to $X$ but it is important not to conflate the two functions, since the arguments to $d$ are word *vectors* and not words per se. That is, $d$ is related to the words in the sample vocabulary only through the vector space defined by the co-occurrence function $X$, and it does not operate directly on the sample vocabulary $\mathcal{V}$. Additionally, the choice of $d$ is not uniquely determined by the choice of $W$. Whether a particular choice of $d$ is appropriate depends primarily on the research goal.

In predictive applications, the most widely-used word association function is *cosine similarity*, the cosine of the central angle formed between two word vectors $a, b \in W$ with respect to the additive identity element of the vector space. This quantity is computed by dividing the Euclidean dot product of the two input vectors by the product of their Euclidean norms:  
$$\cos(a, b) = \frac{a \cdot b}{||a||_2 \ ||b||_2}$$  
This quantity is equivalent to the Pearson product-moment correlation coefficient computed without recentering the input vectors.^[There are various justifications for omitting the centering step. Originally, it was used to analyze word-document co-occurrence frequency matrices, which are sparse by construction since most words are not used in most documents. Cosine similarity can be computed efficiently over pairs of sparse vectors by ignoring indices corresponding to zero values in either vector. However, this trick is not possible when the vectors are centered since the zero values are replaced by the negative of the mean value. See @salton1979mathematics and @turney2010frequency for further discussion.]

Notably, cosine similarity is invariant to rescaling $a$ and $b$ by any two positive constants.^[Strictly speaking, cosine similarity is invariant to any two constants $z_a, z_b$ such that $\text{sign}(z_a) = \text{sign}(z_b)$. This distinction might matter if the signs of the latent vectors are not identified in the underlying model, for example.] Scale invariance is especially desirable in settings where vector norms encode irrelevant information. For example, cosine similarity is a popular ranking heuristic for document retrieval in the word-document co-occurrence frequency regime because document vector norms predominantly reflect the total number of words in the document, and this dimension of variation is rarely of direct interest when attempting to retrieve relevant documents. In the word-word co-occurrence regime, the distribution of word vector norms varies systematically at different levels of marginal word frequency, since this controls the number of unique words that can appear nearby; the exact form of this frequency dependence is determined by the choice of co-occurrence function.^[Cosine similarity also typically yields more stable results than other choices of word association function when studying words that vary substantially in frequency. This data analysis situation occurs frequently in predictive applications, where users tend to be interested primarily in finding informative neighbors of query words: as in any natural language corpus, the words used in search queries tend to be disproportionately high-frequency, but the most informative words in the set of documents being searched are more likely to be low-frequency.]

## Range restriction

In applications of distributional semantic models, it is common for researchers to assert that the cosine similarity between two vectors $a, b \in \mathbb{R}^d$ is bounded in the range $[-1, 1]$. However, this range restriction becomes more complex in the multidimensional comparison regime. In particular, the nominal bounds are misleadingly large for sets of cosine similarities that share input vectors, since if some of the cosines are fixed then the remaining cosines are in general restricted to a subset of the nominal range.

To explain why the range restriction on cosine similarity becomes more stringent in the multidimensional setting, we first show how the range restriction inequality in the univariate case follows from the positive definiteness of the Euclidean inner product $\left< \cdot, \cdot \right> \in \mathbb{R}$ on $\mathbb{R}^d$. Let $a, b \in \mathbb{R}^d$ be two linearly independent word vectors forming the columns of a $d \times 2$ matrix $X$. By construction, the Gram matrix $X^T X$ is a symmetric positive definite matrix with entries equal to the pairwise Euclidean inner products of $a, b$:
$$
X^T X = \begin{bmatrix}
\left<a, a\right> & \left<a, b\right> \\
\left<b, a\right> & \left<b, b\right>
\end{bmatrix}
$$

Next define the leading principal minors of a square matrix to be the determinants of every principal submatrix formed by iteratively deleting the last row and column. For example, the principal minors of $X^T X$ in the $2 \times 2$ case are $\det (\left<a, a\right>) = \left<a, a\right>$ and $\det(X^T X) = \left<a, a\right>\left<b, b\right> - \left<a, b\right>^2$. The following theorem relates the positive definiteness of a symmetric matrix to the signs of its leading principal minors.

#### Theorem (Sylvester's criterion).

A symmetric matrix $M$ is positive definite if and only if all of its leading principal minors are positive.^[For proof see @horn2012matrix, p. 439.]
\hfill\break

Continuing with the univariate case, we obtain $\left<a, a\right> > 0$ and $\left<a, a\right>\left<b, b\right> - \left<a, b\right>^2 > 0$. Rearranging the second inequality yields the familiar form of the range restriction inequality:

$$
\begin{aligned}
\cos^2(a, b) &= \frac{\left<a, b\right>^2}{\left<a, a\right>\left<b, b\right>} < 1 \\
\cos(a, b) &= \frac{\left<a, b\right>}{\left<a, a\right>^{1/2}\left<b, b\right>^{1/2}} \in (-1, 1)
\end{aligned}
$$

We next consider the general multivariate case. Consider two sets of word vectors $A$ and $B$ formed by any two disjoint and deduplicated keyword lists of equal cardinality $k$, and treat these vectors as columns of a $(d \times 2k)$ matrix $X$, where $d$ is the dimension of the word embeddings.^[Note that the ordering of the keyword lists does not matter. Specifically, if we right multiply $X$ by a permutation matrix $M$, then we obtain the Gram matrix $M^T X^TXM$. Since $M$ is an orthogonal matrix, $M^T X^TXM$ is a matrix similarity; all matrices in this equivalence class have identical eigenvalues.] We then form the cosine similarity block matrix $\Theta_{XX}$ partitioned by grouping the vectors by list membership:
$$
\Theta_{XX} = \begin{bmatrix}
\Theta_{AA} & \Theta_{AB} \\
\Theta_{AB}^T & \Theta_{BB}
\end{bmatrix}
$$  

Sylvester's criterion implies that $\det(\Theta_{XX}) = \det(\Theta_{AA})\det(\Theta_{BB} - \Theta_{AB}^T\Theta_{AA}^{-1}\Theta_{AB}) > 0$.^[Note that $\Theta_{AA}, \Theta_{BB}$ are assumed to be nonsingular. For proof of the Schur determinant identity see e.g. @horn2012matrix.] Since $\det(\Theta_{AA})$ must also be positive, it is also the case that $\det(\Theta_{BB} - \Theta_{AB}^T\Theta_{AA}^{-1}\Theta_{AB}) > 0$. Using the notation $X \preccurlyeq Y$ for symmetric $(n \times n)$ matrices $X, Y$ to denote that the matrix $Y - X$ is positive semi-definite, this inequality can be written $\Theta_{AB}^T\Theta_{AA}^{-1}\Theta_{AB} \preccurlyeq \Theta_{BB}$. Keeping in mind that $A$ and $B$ are interchangeable, the blocks of $\Theta_{XX}$ are in general constrained by the following range restriction inequality [@hubert1972note; @olkin1981range; @horn1990cauchy; @bhatia2009positive]:
$$
\begin{aligned}
\Theta_{BB}^{-1/2}\Theta_{AB}^T\Theta_{AA}^{-1}\Theta_{AB}\Theta_{BB}^{-1/2} &\preccurlyeq I_k \\
\end{aligned}
$$

If some of the entries of $\Theta_{XX}$ are fixed then the remaining pairwise cosine similarities are constrained to lie within a subset of their nominal range. This phenomenon complicates the interpretation of the between-set cosine similarities. In the setting we consider in this paper, we fix the within-set cosine submatrices $\Theta_{AA}$ and $\Theta_{BB}$ by selecting keyword lists. This implies that the distribution of the entries of $\Theta_{AB}$ is supported on a more restricted range than the nominal bounds on cosine similarity unless we assume that $\Theta_{AA}$ and $\Theta_{BB}$ equal the identity matrix in expectation.

Higher-order range restrictions have a geometric interpretation. For example, consider the cosine similarity matrix formed by any three word vectors $a \in A$ and $b_1, b_2 \in B$:
$$
\Theta_{XX}
= \left[
\begin{array}{c|cc}
1 & \cos(a, b_1) & \cos(a, b_2)  \\
\hline
\cos(a, b_1) & 1 & \cos(b_1, b_2) \\
\cos(a, b_2) & \cos(b_1, b_2) & 1
\end{array}
\right]
$$
The range restriction inequality implies that the entries of $\Theta_{AB}$ are jointly constrained to lie in the elliptical region $\cos^2(a, b_1) - 2\cos(b_1, b_2)\cos(a, b_1)\cos(a, b_2) + \cos^2(a, b_2) \leq 1 - \cos^2(b_1, b_2)$; the within-set cosine similarity $\cos(b_1, b_2)$ parameterizes the constraint by determining how "elongated" this region is [@jackson1924trigonometry; @hubert1972note; @rousseeuw1994shape].

In the univariate case the constraint always reduces to the familiar unit disk constraint. However, in the two-to-one setting the remaining cosines are bounded within the unit disk only when $\cos(b_1, b_2) = 0$. As the ellipticity of $\Theta_{BB}$ increases --- that is, as we choose more closely related words for the keyword list $B$ --- the range of possible values for the remaining two cosine similarities becomes narrower, so they are constrained to be closer together in value. The attainable range of each remaining cosine depends on the value of the other cosine; in particular, $\cos(a, b_1)$ attains the nominal range of the cosine function only when $\cos(b_1, b_2) = \cos(a, b_2) = 0$.

Additionally, it is possible to construct comparisons in which a subset of the between-set cosine similarities for a given pair of rows/columns in $\Theta_{AB}$ cannot equal zero. In the three-vector case for example, if we assume provisionally that $\cos(a, b_2) = 0$, this implies $\cos^2(a, b_1) \leq 1 - \cos^2(b_1, b_2)$, but if $\cos(b_1, b_2) = 0.8$, then for values $\cos(a, b_1) > 0.6$ it is impossible for $\cos(a, b_2) = 0$.

## Measuring anisotropy

It is useful to be able to measure how much our particular choice of $A$ and $B$ has constrained the entries of $\Theta_{AB}$. However, in most cases the number of words makes it impractical to express the range restriction inequality in terms of the off-diagonal entries of the within-set cosine matrices. To avoid this, we instead measure the level of within-set association using a metric on $\Theta_{AA}$ and $\Theta_{BB}$. Specifically, the level of within-set association for each keyword list can be quantified by measuring how *anisotropic* (directionally non-uniform) the corresponding set of word vectors is. We say that $A$ is isotropic if and only if the cosine similarity matrix $\Theta_{AA} = I_k$.^[We could state this more generally by asserting that $A$ is isotropic if and only if the Gram matrix is a scalar matrix: $A^T A = \phi_AI_k$ for some positive constant scaling factor $\phi_A$, or equivalently $A$ is isotropic if and only if all of its column vectors are mutually orthogonal and equivalent in norm. This condition implies that $\Theta_{AA} = I_k$. In general we will prefer to work with the cosine similarity matrix directly since the scalar anisotropy adjustment is already well-known.] We can then express the total amount of anisotropy induced by $A$ as the distance between $\Theta_{AA}$ and the identity matrix under a suitable metricization of the space of symmetric positive definite matrices. 

The space of symmetric positive definite matrices forms a Riemannian manifold with intrinsic metric $d(X, Y) = ||\text{Log}(X^{-1/2}YX^{-1/2})||_F$, where $||X||_F = \text{tr}(X^T X)^{1/2}$ is the Frobenius norm and $\text{Log}(X)$ denotes the principal matrix logarithm of $X$ [@moakher2005differential; @bhatia2009positive]. Under this metric^[Note that there are other Riemannian geometries on the space of symmetric positive definite matrices; see e.g. @arsigny2007geometric.] the nearest isotropic matrix to $\Theta_{AA}$ is $\hat\phi_A I_k$, where the scaling factor $\hat\phi_A = (\det \Theta_{AA})^{1/k}$ is the geometric mean of the eigenvalues of $\Theta_{AA}$ [@moakher2006symmetric]. We call the geodesic distance between $\Theta_{AA}$ and $\hat\phi_A I_k$ the "isotropy index" of A:  
$$\text{Iso}(A) = ||\text{Log} ((\hat\phi_A I)^{-1/2} \Theta_{AA} (\hat\phi_A I)^{-1/2})||_F$$  

The isotropy index is closely related to Bartlett's test of sphericity, a classic diagnostic test in exploratory factor analysis [@bartlett1951effect]. Specifically, if the correlation matrix $\Theta_{AA}$ for an $d \times k$ matrix $A$ is the identity matrix, then $f(A) = -\log(\det(\Theta_{AA}))(d - 1 - (2k + 5)/6)$ is approximately distributed as $\chi^2(k(k-1)/2)$, where $d$ is the number of observations and $k$ is the number of variables. The test statistic very closely approximates $\text{Iso}(A)$, where $d$ is instead the dimension of the ambient embedding space and $k$ is the cardinality of the set of word vectors; the approximation becomes less exact for larger values of $k$.

Note that the nearest isotropic matrix is not general a cosine similarity matrix, since the only isotropic cosine similarity matrix is the identity matrix and in general $\hat\phi_A < 1$. We might then wonder, why not measure anisotropy as the geodesic distance between $I_k$ and $\Theta_{AA}$ directly? Although $I_k$ and $\Theta_{AA}$ are both cosine similarity matrices, elements between them along the symmetric positive definite manifold are not constrained in this way, and whether a given path is geodesic depends on how we project its elements onto the set of cosine similarity matrices. We therefore need to specify a geometry for the space of cosine similarity matrices. @david2019riemannian characterize the space of correlation matrices as an isometric quotient submanifold of the space of positive definite matrices by the diagonal matrices with strictly positive diagonal entries. Under this geometry any isotropic matrix is in the equivalence class of the identity matrix; since $\hat\phi_A I_k$ is the nearest isotropic matrix, $\text{Iso}(A)$ measures the geodesic distance to isotropy.


# Measuring multidimensional semantic association

We next compare two strategies for measuring the association between sets of word vectors interact with the range restriction inequality. MCS-based methods target arithmetic means of the between-set cosine similarities (e.g., the cosine similarity between their centroids). GCD instead targets the cosine similarity between the subspaces spanned by the two sets of word vectors. Both approaches are analyzed below as functions of the blocks of $\Theta_{XX}$, the matrix of cosine similarities between pairs of elements across $A$ and $B$. As before, we assume that the lists used to form $A$ and $B$ are disjoint and deduplicated.

## Mean cosine similarity

Many proposed methods for multidimensional semantic measurement target an arithmetic mean of the entries of $\Theta_{AB}$. Let $\eta$ be an $k \times k$ matrix of weights with entries $\eta_{ij} \in \mathbb{R}$. All MCS-based measures are equivalent to the Frobenius inner product of the between-cosine matrix with $\eta$:
$$d_{\text{MCS}}(A, B) = \sum_{i=1}^k\sum_{j=1}^k \eta_{ij} d_{\cos}(A_i, B_j) = \text{tr}(\eta^T \Theta_{AB})$$

The validity of this quantity as a measure of association between the target concepts depends on two assumptions.

First, we must assume that the selected keyword lists identify the targeted latent variation. We do not characterize this assumption in detail.

Second, we must assume that the keyword selection mechanism induces sets of word vectors that are isotropic in expectation (i.e., $E[\text{Iso}(A)] = E[\text{Iso}(B)] = 0$). If the second assumption holds, the distribution of $\Theta_{AB}$ does not depend on $\Theta_{AA}, \Theta_{BB}$. In this situation the arithmetic mean estimates the expected cosine similarity between any pair of words drawn uniformly at random from each list.

Many proposals for embedding-based measures of semantic association are MCS-based methods with a particular choice of $\eta$. For example, when the cosine similarities are weighted equally, $\eta$ is the ones matrix times some normalizing scalar. The MCS-based family of measures includes methods that first compute centroids or similar plug-in linear combinations of word vectors. It also includes any linear regression model with a cosine-valued outcome; in this setting the weights $\eta_{ij} = p(d_{\cos}(A_i, B_j)|X_{ij})$ are estimates of the probability of each cosine similarity conditional on covariates, and the MCS quantity is then interpreted as a conditional expectation. We next discuss a few illustrative cases.

### Word embedding association tests

Many centroid-based MCS quantities are modeled after the Word Embedding Association Test method ["WEAT"; see @caliskan2017semantics]. WEAT models target the two-way difference in mean pairwise cosine similarity between sets of word vectors corresponding to paired keyword lists. Formally, let $A, B; C, D$ be the paired subsets of row vectors in $W$ induced by the choice of keyword lists. For example, a typical WEAT problem might involve measuring the differential association of $A$ = Maleness and $B$ = Femaleness with $C$ = Pleasantness and $D$ = Unpleasantness. The WEAT score for the differential association between Pleasantness/Unpleasantness and Maleness/Femaleness is the two-way difference in arithmetic mean cosine similarity between pairs of vectors across $A, B$ and $C, D$, i.e.:  
$$
\begin{aligned}
\text{WEAT}(A, B; C, D) &= \frac{1}{k}\sum_i^k\sum_j^k\Big(d_{\cos}(A_i,C_j) - d_{\cos}(A_i,D_j) + d_{\cos}(B_i,D_j) - d_{\cos}(B_i,C_j)\Big)
\end{aligned}
$$  
So WEAT consists of a comparison between four uniformly weighted MCS quantities (i.e., $\eta$ is a constant matrix with entries $1/k$). Note that the problem is symmetric up to a sign change if $A, B$; $C, D$; or $\{A, B\}$; $\{C, D\}$ are swapped.

### Semantic centroid methods

A popular modification to the WEAT setup targets the cosine similarity between a word vector and a semantic "axis", "dimension", or "direction" obtained by subtracting the centroids of the two word vector sets, i.e. $\cos(\bar{A} - \bar{B}, w)$ [@kozlowski2019geometry; @stoltz2023tool]. Models of this type were developed following the hypothesis that arithmetic on word vectors acts as a kind of semantic algebra so that adding and subtracting word vectors can be used to solve analogy problems [@mikolov2013linguistic], but there is some debate about how well this measurement strategy works [see for example @church2017word2vec].

The cosine similarity $\cos(\bar{A} - \bar{B}, w)$ is an MCS-based quantity:  
$$
\begin{aligned}
d_{\cos}(\bar{A} - \bar{B}, w) &= \frac{\bar{A} \cdot w}{||\bar{A} - \bar{B}|| \ ||w||} - \frac{\bar{B} \cdot w}{||\bar{A} - \bar{B}|| \ ||w||} \\
&= \sum_i^k {\Large (} \frac{A_i \cdot w}{||\bar{A} - \bar{B}|| \ ||w||} - \frac{B_i \cdot w}{||\bar{A} - \bar{B}|| \ ||w||} {\Large )}\\
&= \sum_i^k {\Large (}\frac{||A_i||}{||A_i||} \frac{A_i \cdot w}{||\bar{A} - \bar{B}|| \ ||w||} - \frac{||B_i||}{||B_i||} \frac{B_i \cdot w}{||\bar{A} - \bar{B}|| \ ||w||} {\Large )}\\
&= \sum_i^k  \frac{||A_i||}{||\bar{A} - \bar{B}||} d_{\cos}(A_i, w) - \sum_i^k\frac{||B_i||}{||\bar{A} - \bar{B}||} d_{\cos}(B_i, w)\\
\end{aligned}
$$
Each of the cosine similarities is weighted by the ratio of the norm of each vector in $A$ or $B$ to the norm of $\bar{A} - \bar{B}$. Since vector addition is commutative, we can view $\bar{A} - \bar{B}$ as either the difference between the centroids or the centroid of subtracted vector pairs (i.e. the centroid of the set of vectors $A - B$) as long as $A$ and $B$ are identical in cardinality and do not contain any duplicate word vectors. If either set contains repeated vectors, this is analogous to upweighting the duplicated word vector.

The centroid-centroid cosine $d_{\cos}(\bar{A} - \bar{B}, \bar{C} - \bar{D})$ is analogous to the WEAT quantity. It is numerically equivalent to a weighted two-way difference in mean cosine similarity, with weight $\eta_{ij}$ equal to the ratio of the product of the norms of the $ij$-th vector pair to the product of the norms of the semantic axes. This quantity can also be written as a weighted mean dot product between elements of (e.g.) $A$ and $C$:  
$$
\begin{aligned}
d_{\cos}(\bar{A} - \bar{B}, \bar{C} - \bar{D}) &= h(\bar{A}, \bar{C}) - h(\bar{A}, \bar{D}) + h(\bar{B}, \bar{D}) - h(\bar{B}, \bar{C}) \\
h(\bar{A}, \bar{C}) &= \sum_i^k \sum_j^k \frac{||A_i||||C_j||}{||\bar{A} - \bar{B}|| \ ||\bar{C} - \bar{D}||} d_{\cos}(A_i, C_j) \\
&= \frac{1}{||\bar{A} - \bar{B}|| \ ||\bar{C} - \bar{D}||} \sum_i^k \sum_j^k (A_i \cdot C_j)
\end{aligned}
$$


## Generalized coefficient of determination

MCS-based methods represent keyword lists as arbitrary linear combinations of the corresponding sets of word vectors. In lieu of fixing the coefficients on these combinations exactly, a more general representation can be obtained by forming the equivalence class of *any* linear combination of the word vectors. This equivalence class is the linear subspace of $\mathbb{R}^d$ spanned by each set of word vectors; it is uniquely identified with the orthogonal projection matrix $P_A = U_AU_A^T$, where $A = U_A S_A V_A^T$ is the singular value decomposition of $A$. By construction, $P_A$ is invariant to $V_A, S_A$, so it does not depend on the covariance of the columns of $A$.

The generalized coefficient of determination or GCD [@rao1979general; @yanai2011projection; @ramsay1984matrix; @meyer2000matrix; @borg2005modern; @horn2012matrix] measures the multivariate association between $A$ and $B$ as the cosine similarity between $P_A$ and $P_B$ viewed as vectors in $\mathbb{R}^{d \times d}$:  
$$
\begin{aligned}
d_{\text{GCD}}(A, B) 
= \frac{\left<P_A,P_B\right>_F}{||P_A||_F||P_B||_F}
= \frac{\text{tr}(P_A P_B)}{\sqrt{k_Ak_B}}
\end{aligned}
$$
Since $A$ is assumed to be full rank, the denominator norms are fixed exactly by choosing the number of keywords in each list, and if $k_A = k_B$ then the normalizing constant is just $k^{-1}$.

$d_{\text{GCD}}$ is closely related canonical correlation analysis (CCA; CITE Hotelling 1936). The maximal canonical correlation is the cosine of the minimal angle between any pair of linear combinations of vectors in $A$ and $B$; these linear combinations are called the first canonical variates of $A$ and $B$ with respect to one-another. The subsequent canonical correlations measure the subsequent minimal angles between the remaining canonical variates subject to the constraint that successive pairs of canonical variates are orthogonal to every other canonical variate pair for $A$ and $B$, so that the two sets of canonical variates form the pair of orthogonal bases for $A$ and $B$ with minimal angular separation. $d_{\text{GCD}}$ is the sum of the squared canonical correlations between $A$ and $B$ normalized by the geometric mean of the cardinalities of $A$ and $B$. This connection makes computing $d_\text{GCD}$ simple and convenient in any statistical software with a CCA implementation.^[In R for example, where `A, B` are matrices with columns equal to the word vectors in each set, $d_\text{GCD}$ is equal to `sum(cancor(A, B, F, F)$cor^2)/sqrt(ncol(A) * ncol(B))`.]

The standard definition of $d_\text{GCD}$ assumes that the input data matrices $A$ and $B$ are column-centered, whereas typically cosine similarity is defined as the *non-recentered* correlation between the input vectors. When applying $d_{\text{GCD}}$ in the semantic analysis setting, we do not recenter $A$ and $B$ on their local centroids. In practice centering does not affect the measure very much since the centering projection of a matrix does not change the order of its eigenvectors [@cadima2009relationships].

$d_{\text{GCD}}$ naturally extends cosine similarity to multidimensional comparisons. This can be seen by rewriting the inner product $\text{tr}(P_AP_B)$ in terms of the range restriction inequality. Fix $k_A = k_B$, and let $Q_A, Q_B$ be the respective inverses of two diagonal matrices with nonzero entries equal to the Euclidean norms of each column of $A, B$ (i.e., $[Q_A]_{ii} = 1/||A_i||_2$). Applying the cyclic permutation invariance and convexity of the trace:
$$
\begin{aligned}
\text{tr}(P_A P_B) &= \text{tr}(A(A^T A)^{-1}A^T B(B^T B)^{-1} B^T) \\
&= \text{tr}(B^T A(A^T A)^{-1}A^T B(B^T B)^{-1}) \\
&= \text{tr}(Q_BB^T AQ_A(Q_AA^T AQ_A)^{-1}Q_AA^T BQ_B(Q_BB^T BQ_B)^{-1}) \\
&= \text{tr}(\Theta^T_{AB}\Theta^{-1}_{AA}\Theta_{AB}\Theta^{-1}_{BB}) \\
&= \text{tr}(\Theta^{-1/2}_{BB}\Theta^T_{AB}\Theta^{-1}_{AA}\Theta_{AB}\Theta^{-1/2}_{BB})
\end{aligned}
$$
It follows that $\text{tr}(P_A P_B) \leq \text{tr}(I_k) = k$, so $k^{-1}\text{tr}(P_A P_B) \in [0, 1]$.



## Anisotropic distortion in MCS

Previously, we established that any mean cosine similarity can be written as $\text{tr}(\eta \Theta_{AB})$, the Frobenius inner product of the between-cosine matrix with a weight matrix $\eta$. We next show that $d_\text{GCD}$ can be viewed as the *unique* choice of $\eta$ that makes $d_\text{MCS}$ invariant to the anisotropy of $A$ and $B$.

We first establish that $d_{\text{GCD}}$ is in fact invariant to any right nonsingular transformation of $A$ and/or $B$. Let $M_A, M_B$ be two arbitrary $k \times k$ nonsingular linear transformations of the ranges of $A$ and $B$. Denote the "original" matrices of word vectors $\tilde{A}, \tilde{B}$, and define the equivalence classes $A = \tilde{A}M_A$ and $B = \tilde{B}M_B$. All such matrices yield the projection matrices $P_A, P_B$ irrespective of the choice of $M_A, M_B$:

$$
\begin{aligned}
P_{AM_A} &= AM_A(M_A^TA^TAM_A)^{-1}M_A^TA^T \\
&= AM_AM_A^{-1}(A^T A)^{-1}M_A^{-T}M_A^TA^T \\
&= A(A^TA)^{-1}A^T = P_A. \\
\end{aligned}
$$
Hence $A$ and $B$ may each exhibit any level of anisotropy; further right nonsingular transformations do not affect the left singular vectors of $\tilde{A}$.^[The embedding space may also be left rotated *globally* by any $d \times d$ orthogonal transformation of the embedding space. If $R$ is any orthogonal matrix then $P_{RA} = RA(A^TR^TRA)^{-1}A^TR^T = RP_AR^T$. By symmetric application to $B$ the Frobenius inner product of the projection matrices becomes $\text{tr}(RP_AR^TRP_BR^T)$, and by the similarity invariance of the trace this quantity is equivalent to $\text{tr}(P_AP_B)$.] This crucial property makes $d_{\text{GCD}}$ a superior choice for measuring between-set word association when the input keyword lists are selected non-randomly and the level of anisotropy is not zero in expectation for the chosen model.

We next show that there is no other choice of weights satisfying this invariance property. We begin by writing the MCS quantity $d_{\text{MCS}}(A, B)$ in terms of the singular value decompositions of $A, B$:  
$$
\begin{aligned}
d_{\text{MCS}}(A, B) &= \text{tr}(\eta^T {\Theta}_{AB}) \\
&= \text{tr}(\eta^T {Q}_A A^T B {Q}_B) \\
&= \text{tr}(B {Q}_B \eta^T {Q}_A A^T) \\
&= \text{tr}(U_B S_B V^T_B {Q}_B \eta^T {Q}_A V_A S_A U^T_A) \\
\end{aligned}
$$  
As before, the matrices ${Q}_A, {Q}_B$ are the diagonal matrices that rescale the cross-Gram matrix into the corresponding between-cosine matrix (i.e., $\Theta_{AB} = Q_A\Sigma_{AB}Q_B$).

Whenever the matrix $S_B V^T_B {Q}_B \eta^T {Q}_A V_A S_A$ does not equal the identity matrix in expectation, $d_{\text{MCS}}$ depends on how anisotropic $A$ and $B$ unless we choose the weight matrix $\eta$ to ensure that the scale and covariance factors $S_A$, $S_B$, $V_A$, $V_B$, ${Q}_A$, and ${Q}_B$ drop out of the matrix product exactly. A choice of weights that satisfies this requirement is $\hat\eta_\text{GCD}^T = {Q}_B^{-1} V_B^{-T} S_B^{-1} U^T_B U_A S_A^{-1} V_A^{-1} {Q}_A^{-1}$. Substituting this matrix for $\eta^T$ yields $d_\text{GCD}$:
$$
\begin{aligned}
&= \text{tr}(U_B S_B V^T_B {Q}_B \hat\eta_\text{GCD}^T {Q}_A V_A S_A U^T_A) \\
&= \text{tr}(U_B S_B V^T_B {Q}_B {Q}_B^{-1} V_B^{-T} S_B^{-1} U^T_B U_A S_A^{-1} V_A^{-1} {Q}_A^{-1} {Q}_A V_A S_A U^T_A) \\
&= \text{tr}(U_B U^T_B U_A U^T_A). \\
\end{aligned}
$$

Notably, an alternative weight matrix $\eta_* = {Q}_B^{-1} V_B^{-T} S_B^{-1} S_A^{-1} V_A^{-1} {Q}_A^{-1}$ could be obtained by omitting the central $U^T_B U_A$ matrix product, and this choice of weights also makes MCS invariant to right nonsingular transformations of the input vectors. This yields a different measure of association based on $\text{tr}(U_B U^T_A)$, the Frobenius inner product between the left singular vectors $U_A, U_B$ of $A$ and $B$. However, the left singular vectors of $A$ and $B$ are only unique up to an arbitrary choice of sign, so this quantity is not reflection invariant. We can restore reflection invariance by instead considering the subspace spanned by the singular vectors. The orthogonal projection $U_A U_A^T$ (e.g.) uniquely identifies this subspace; we can choose any orthogonal reflection matrix $Z_A$ (a diagonal matrix with nonzero entries in $\{-1, 1\}$) applied on the right to $U_A$, and this becomes the identity metric when forming the projection matrix.

Since $\hat\eta_\text{GCD}$ is the only choice of weight matrix that yields exact anisotropy invariance, it follows that $d_\text{MCS}$ is distorted under other choices of weights. To characterize the amount of distortion numerically, we next evaluate $d_\text{MCS}$ and $d_\text{GCD}$ in a set of example association problems.


# Effects of anisotropy on multidimensional semantic association measures

This section compares the performance of MCS and GCD under varying levels of anisotropic input vector selection in a realistic setting where the targeted subspaces are not known a priori. To do this we use a collection of keyword lists from the applied word embeddings literature along with a set of word embeddings trained on a widely-used text corpus.

We obtain word vectors using word co-occurrences in English Wikipedia article text as archived in September 2023. The text is preprocessed to remove punctuation and symbols, and all words are lowercased. Words occurring fewer than 50 times are removed from the text to manage the dimension of the sample vocabulary. Word co-occurrences are counted up to 50 words to the left and right each word. We then use GloVe to obtain a vector in $\mathbb{R}^{300}$ for each keyword [@pennington2014glove]; we use only the $W$ vectors to represent each word. GloVe downweights prediction errors for co-occurrence counts below a user-specified threshold during estimation; here the low-frequency threshold hyperparameter is set to 200.

To construct example conceptual association problems, we use a set of 37 keyword lists introduced in prior work on word embeddings in the social sciences [@caliskan2017semantics; @kozlowski2019geometry]. This set of keyword lists is meant to be representative of the keyword lists used by researchers in the social sciences. The substantive associations these comparisons imply may or may not be theoretically interesting; the constructed comparisons are meant to be illustrative of the mathematical structure of the problem rather than diagnostic for particular topics of semantic analysis. Minor modifications to the keywords (i.e., spelling changes) are necessary to ensure support in this corpus; the full set of keyword lists and an enumeration of modifications is available in the Appendix. As noted previously, since characterizing subspace estimation error due to keyword selection is beyond the scope of this paper, all results treat the keyword lists as fixed; see Discussion for further remarks on this aspect of the problem.


## Keyword selection induces anisotropy

Figure \ref{fig1} compares the observed isotropy index $\text{Iso}(A)$ for each keyword list (red point) to the null distribution obtained by repeatedly sampling keyword lists of equal cardinality uniformly at random from the sample vocabulary (black bar covers 99% of generated samples).^[Other random keyword sampling strategies are possible. For example, we could instead sample only words occurring not more/less than the maximum/minimum frequency keyword in the current keyword lists, so that the random lists are similar to the observed lists in overall frequency.] The lists are ordered along the Y-axis by cardinality, and the X-axis displays the value of the isotropy index. Observed keyword lists are in general markedly more anisotropic than we would expect to observe by randomly drawing keyword lists of identical cardinality. Notably, the expected anisotropy for random keyword lists is nonzero and increasing in the number of keywords, so even random sets of word vectors yield much more anisotropic subspace representations than we would observe by (e.g.) randomly sampling vectors uniformly in $\mathbb{R}^{300}$.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.9\textwidth]{/Users/akindel/code/cosine/figures/figure1_keyword_anisotropy.pdf}
\caption{Isotropy indices for selected keyword lists (see Appendix). Larger values indicate that the range restriction on sets of pairwise cosine similarities involving this keyword list is more stringent. The observed keyword lists (red points) are markedly more anisotropic than randomly sampled keyword lists of identical cardinality (black bar indicates 99\% of 2000 samples). Randomly sampled sets of word vectors are themselves more anisotropic than we would expect to observe by sampling the same number of random vectors distributed uniformly with respect to direction (i.e., isotropically).}
\label{fig1}
\end{center}
\end{figure}


## MCS distortion under increasing anisotropy

To obtain *more* anisotropic representations for $A$ and $B$ we generate two order $k$ nonsingular matrices $S, T$ and compute $A=AS, B=BT$. The transformations are of the form $I_k + X$, where $I_k$ is the order-$k$ identity matrix and the entries of $X$ are i.i.d. Gaussian draws with mean zero and variance $\theta$. Choosing regularized nonsingular transformations of this form allows $\theta$ to parameterize the size of the transformation such that the transformation converges to the identity matrix as $\theta \rightarrow 0$. As $\theta$ increases, the expected value of the isotropy index converges to $\sqrt{d}$ where $d$ is the dimension of the ambient vector space, since it is invariant to non-uniformity in the norms of the vectors in $A$.

Figure \ref{fig2} shows that the range of MCS is reduced as $A$ and $B$ become increasingly anisotropic under simulated perturbations. Each point corresponds to the average values of $d_\text{MCS}(A, B)$ and $\text{Iso}(A) + \text{Iso}(B)$ for a single pair of keyword lists over 100 simulated transformations of the size indicated by the color of the point. As the perturbations become large, MCS approaches zero in expectation.

Figure \ref{fig3} displays this phenomenon for two example association problems; here each point represents a single simulated perturbation, and the starting values of $d_\text{MCS}(A, B)$ and $\text{Iso}(A) + \text{Iso}(B)$ indicated by the dashed and dotted lines respectively. Note that it is possible to obtain values of MCS that are either larger or smaller than the observed value even under anisotropy-increasing transformations, so the distortion in the quantity due to anisotropy is not conservative in general. This phenomenon occurs more commonly for smaller transformations but is possible even when the change in the amount of anisotropy is relatively large.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.9\textwidth]{/Users/akindel/code/cosine/figures/figure2_numerical_perturbation.pdf}
\caption {Change in MCS distribution under increasingly anisotropic transformations.}
\label{fig2}
\end{center}
\end{figure}


\begin{figure}[H]
\begin{center}
\includegraphics[width=\textwidth]{/Users/akindel/code/cosine/figures/figure3_perturbed_examples.pdf}
\caption {Anisotropic distortion in example association problems. In general MCS quantities shrink to zero as more anisotropic transformations are applied, but it is also possible to observe higher values even when the amount of anisotropy is higher, especially when the transformation is not very large.}
\label{fig3}
\end{center}
\end{figure}



## MCS distortion under decreasing anisotropy

To obtain *less* anisotropic representations for $A$ and $B$, it is inefficient to randomly generate nonsingular perturbations using the above method, since most directions in the space of SPD matrices are anisotropy-increasing. To examine how anisotropy reduction affects MCS-based measures, we instead employ geodesic interpolation to move the within-set cosine similarity matrices $\Theta_{AA}, \Theta_{BB}$ toward the identity matrix directly. We obtain elements of the geodesics between $\Theta_{AA}, \Theta_{BB}$ and the identity matrix on the manifold of cosine similarity matrices using the projective method of @david2019riemannian. We first construct the geodesic between each cosine similarity matrix and its nearest isotropic matrix on the ambient manifold of positive definite matrices as described above; we then project elements of this geodesic onto the space of cosine similarity matrices. Specifically, where $\pi(X) = \text{Diag}(X)^{-1/2} X \text{ Diag}(X)^{-1/2}$ is the operator projecting symmetric positive definite matrices onto the submanifold of cosine similarity matrices, the element of the geodesic located $100t\%$ of the way from $\Theta_{AA}$ to $\hat\phi_A I$ is given by:
$$\tilde\Theta_{AA}(t) = \pi(\Theta_{AA}^{1/2} \text{ Exp}[ t \text{ Log} (\Theta_{AA}^{-1/2} (\hat\phi_A I) \Theta_{AA}^{-1/2})]\Theta_{AA}^{1/2})$$

This procedure allows us to determine the value of $t$ beyond which $\Theta_{AA}$ and $\Theta_{BB}$ cannot be made more isotropic while preserving the positive definiteness of the supervening cosine similarity matrix $\Theta_{XX}$. For interpolants closer to the identity matrix than this point, the observed between-set cosine similarity matrix $\Theta_{AB}$ is geometrically impossible. Interpretations of MCS-based quantities that assume $A$ and $B$ are isotropic are invalid beyond this breakdown point.

Figure \ref{fig4} displays the grid of breakdown points obtained by simultaneously reducing the anisotropy of $\Theta_{AA}$ and $\Theta_{BB}$ for all possible keyword list pairs. Notably, $\Theta_{AB}$ is geometrically consistent with exact isotropy in only 24.9% (166/666) of the comparisons tested. If the set of keyword lists is viewed as an i.i.d. random sample of keyword lists for *any* target concept, we would therefore expect 3 in 4 MCS-based comparisons to be invalid.

We can also extend this procedure to smoothly interpolate between MCS-based quantities and GCD. To do this, for $t \in (0, 1)$ we interpolate linearly between $\Theta_{AB}^T$ and the matrix of normalizing weights implied by the choice of mean, and we simultaneously interpolate the within-set cosine matrices geodesically using the same parameter. We then compute GCD using the interpolants instead of the corresponding observed matrices.

Figure \ref{fig5} displays two example interpolation curves between GCD (dashed line; $t=0$) and uniformly weighted MCS (dotted line; $t=1$). The color of each point indicates whether $\Theta_{XX}$ is consistent with the geodesic interpolation of the within-set cosine matrices; note that this breakdown point depends only on the within-set cosine interpolants as previously described. The interpolation curve is in general nonlinear, and it is nonmonotonic in 92.6% (617/666) of the comparisons tested. Consequently, while the values of MCS and GCD may appear close in an arithmetic sense, this comparison can dramatically understate the total amount of anisotropic distortion in MCS.

\begin{figure}[H]
\begin{center}
\includegraphics[width=\textwidth]{/Users/akindel/code/cosine/figures/figure4_geodesic_interpolation_grid.pdf}
\caption {Breakdown points for keyword list pairs under smooth geodesic anisotropy reduction. Darker colors indicate larger distances between the breakdown point and exact isotropy. Values are rounded to the nearest increment of 0.02. For approximately 75\% of observed keyword list pairs, it would be geometrically impossible for the between-set cosine similarity matrix to occur if the within-set cosine similarity matrices were exactly isotropic.}
\label{fig4}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.9\textwidth]{/Users/akindel/code/cosine/figures/figure5_geodesic_interpolation_examples.pdf}
\caption {Example interpolation curves between MCS and GCD. Interpolants are displayed in increments of 0.02. The breakdown point lies between the last green interpolant and the first red interpolant. The interpolation curve is nonmonotonic in 92.6\% of keyword list pairs, so the arithmetic difference between MCS and GCD generally understates how different the underlying representations are.}
\label{fig5}
\end{center}
\end{figure}


# Discussion

Using GCD in place of other MCS-based quantities addresses one important source of variation in multidimensional semantic association problems: the level of association within each set of keywords selected into the analysis. Empirically, we observe that keyword selection induces sets of word vectors that exhibit markedly higher anisotropy than we would expect if an equal number of word vectors were drawn uniformly at random from the vocabulary. Since in this setting the level of within-set association is attributable entirely to the selection mechanism and not the variation of interest, it is preferable to use a measure of association that is anisotropy-invariant. GCD is the unique element of the class of MCS-based quantities that satisfies this requirement. However, using GCD does not guarantee that the chosen subspace is in fact a good representation of the intended semantic variation. Hence an important limitation of this paper is that it does not systematically characterize the keyword selection mechanism. We conclude the paper by commenting briefly on this aspect of multidimensional semantic analysis.

The main practical challenge associated with keyword selection is that the concepts that social scientists tend to be interested in measuring are almost never identified with a single fixed keyword list. To validate keyword lists, researchers tend to rely on some evidence that they are a priori admissible for a known target concept. For example, researchers commonly derive lists from prior experiments testing the association between the keywords and the concept label; by finding words grouped under the concept label in a thesaurus; or by generating lists using their own sense of word association. These validation steps are reasonable starting points for extracting the targeted variation, but they provide little hope of ruling out the vast pool of implicit alternative keyword lists that this process generates. Even if some subset of the possible keywords for a given concept can be selected provisionally, the number of related-but-inessential words that could be added or removed from this list implies a large set of alternative keyword lists that are equally admissible in a qualitative sense. When it is not possible to identify a single "best" keyword list and the chosen keyword list is not plausibly a random draw from this set, the choice of keywords is an essentially arbitrary decision. We hope to see future work that directly characterizes how much our estimates of semantic association depend on the choice of input words.

More generally, the results in this paper reinforce the importance of thinking carefully about the logic of *construct validation* for text-based measurement models in the social sciences. When using a model to measure a hypothesized kind of latent variation in empirical data, it is essential to evaluate how well the target quantity extracted from the model represents the variation of interest [@lundberg2021your]. This step in social measurement is a crucial point of departure from purely predictive modeling, where the best model is chosen regardless of whether its internal structure embeds any particular theoretical construct. Although it is coherent to prefer theoretical constructs that can be used to make strong predictions (see e.g. @watts2014common), it does not follow that the most predictive models automatically correspond to the specific theoretical constructs we happen to have in mind when we specify them.

The intensifying use of machine learning methods for social measurement may have generated some confusion around this point. In the machine learning literature, models are evaluated primarily in terms of their performance on predictive "benchmark" tasks. Researchers adapting these models to social science tasks have adopted this validation approach; for example, in most papers proposing MCS-based methods, the primary form of validation involves checking for high correlation between the between-set cosine similarities and human ratings of word association [@kozlowski2019geometry; @boutyline2023forging]. Yet strong performance on these predictive tasks provides very little assurance that downstream quantities based on the between-set cosines are in fact measurements of the variation the researcher has in mind. As researchers continue to adapt models optimized for prediction to research questions requiring representations of specific theoretical constructs, it will remain important to choose quantities of interest that isolate the theorized variation as precisely and completely as possible. 



\pagebreak

## Funding {-}

This work was supported by the Agence Nationale de la Recherche under the France 2030 program (ANR-22-EXES-0014).

## Acknowledgements {-}

I am grateful to audiences at Princeton University, Sciences Po, and Ecole Polytechnique for helpful feedback and conversation on prior versions of this paper. I also wish to thank Anna Berg, Dahyun Choi, Tim Hansen, Jeff Lockhart, Armin Pournaki, Benjamin Rohr, Nicolas Torres-Echevarry, and Fred Wherry for critical commentary on prior drafts; Ning Zhou and Nathan West for introducing me to an aleatoric representation of multidimensional word association problems; and Brandon Stewart for extensive guidance over the lifespan of this project.

## Data availability {-}

All data and code files needed to reproduce the results in the paper can be found at https://github.com/atkindel/cosine_similarity. A permanent copy of the replication materials will be archived at Harvard Dataverse upon publication.

## Competing interests {-}

None.

\pagebreak


## References {-}

<div id="refs"></div>



```{r setup, echo=F, message=F, warning=F}
library(tidyverse)
library(magrittr)
library(pracma)
library(expm)

# Plot theming
theme_self <- function(base_size = 14)
{
  txt <- element_text(size = 14, colour = "black", face = "plain")
  bold_txt <- element_text(size = 14, colour = "black", face = "bold")

  theme_bw(base_size = base_size) +
  theme(
    legend.key = element_blank(), 
    strip.background = element_blank(), 

    text = txt, 
    plot.title = txt, 

    axis.title = bold_txt, 
    axis.text = txt, 

    legend.title = bold_txt, 
    legend.text = txt ) 
}

# Document settings
theme_set(theme_bw())
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE,  # don't pollute the PDF with error messages
                      cache=TRUE, cache.lazy=FALSE, eval=F,  # set eval=T to rerun the analysis
                      dpi=300, fig.height=9)  # figure settings
set.seed(271828)
```

```{r functions, eval=F}
# check associates of a given word w using embedding matrix M
check_neighbors <- function(w, M) {
  wv <- M[,w]
  w.cs <- sapply(1:ncol(M), function(i) coop::cosine(M[,i], wv))
  data.frame(word=colnames(M), cossim = w.cs)
}

# heatmap plot
q.heatmap <- function(S) reshape2::melt(S) %>% ggplot(aes(x=Var1, y=Var2, fill=value)) + geom_tile() + scale_fill_viridis_c()

# projection matrix onto column space of A
projector <- function(A) A %*% solve(t(A) %*% A) %*% t(A)
projector.fast <- function(A) tcrossprod(qr.Q(qr(A)))

# canonical correlations between A and B (Bjorck and Golub 1973)
fast.cca <- function(A, B) {
    AQ <- qr.Q(qr(A))
    BQ <- qr.Q(qr(B))
    svd(crossprod(AQ, BQ), 0, 0)$d
}

# Yanai's GCD: arithmetic mean squared canonical correlation, or cos(P(A), P(B))
# gcd <- function(A, B) mean(fast.cca(A, B)^2)  # this works only when ka=kb...!
gcd <- function(A, B) {
  kA <- ncol(A)
  kB <- ncol(B)
  sum(fast.cca(A, B)^2)/sqrt(kA*kB)
}
gcd.alt <- function(A, B) coop::cosine(c(projector.fast(A), c(projector.fast(B))))

# quadratically regularized GCD
rgcd <- function(A, B, l) {
  ka <- ncol(A)
  kb <- ncol(B)
  Pa <- A %*% solve(t(A) %*% A + diag(l, ka)) %*% t(A)
  Pb <- B %*% solve(t(B) %*% B + diag(l, kb)) %*% t(B)
  Trace(Pa %*% Pb) / sqrt(ka * kb)
}

# unique anisotropy invariant weight matrix for weighted mean cosine similarity
gcd.weight.matrix <- function(A, B) {
  As <- svd(A)
  Bs <- svd(B)
  t(diag(apply(B, 2, norm, "2")) %*% Bs$v %*% diag(1/Bs$d) %*% t(Bs$u) %*% As$u %*% diag(1/As$d) %*% t(As$v) %*% diag(apply(A, 2, norm, "2")))
}

# examine change in GCD when dimensions are double-scaled one at a time
# this is a very heuristic way to understand which dimensions are influencing
#  in the canonical correlation structure between A and B the most
stretch.gcd <- function(A, B, w=2) {
  sapply(1:300, function(i) gcd(sweep(mcv, 1, c(rep(1, i - 1), w, rep(1, 301 - i - 1)), "*")[,A], sweep(mcv, 1, c(rep(1, i - 1), w, rep(1, 301 - i - 1)), "*")[,B]))
}

# GCD after sphering/"de-stretching" embedding dimensions
sphere.gcd <- function(A, B) {
  mcv.norm <- sweep(mcv, 1, apply(mcv,1,norm,"2"), "/")
  gcd(mcv.norm[,A], mcv.norm[,B])
}

# fast leverages: diagonal entries of proj(A)
# this is also helpful for characterizing dimensionwise influence
leverage.fast <- function(A) {
  AQ <- qr.Q(qr(A))
  rowSums(AQ^2)
}

# random orthogonal matrix in n dimensions
# sample uniformly with respect to haar measure
# could also choose nearest element of SO(n)
rotator <- function(n) {
  M <- matrix(rnorm(n*n), n, n)
  Mqr <- qr(M)
  Wd <- diag(qr.R(Mqr))
  W <- diag(Wd/abs(Wd))
  R <- qr.Q(Mqr) %*% W
  R
  # if(det(R) < 0) {
  #   Ms <- svd(M)
  #   Ms$u[,ncol(Ms$u)] <- -1 * Ms$u[,ncol(Ms$u)]
  #   M <- Ms$u %*% diag(Ms$d) %*% t(Ms$v)
  # }
  # M
}

# *small* random proper rotation matrix in n dimensions
# e parameter controls the amount of rotation
rotator.small <- function(n, e=0.01) {
  M <- qr.Q(qr(diag(1, n) + matrix(rnorm(n^2, 0, e), n, n)))
  Mqr <- qr(M)
  Wd <- diag(qr.R(Mqr))
  W <- diag(Wd/abs(Wd))
  R <- qr.Q(Mqr) %*% W
  R
  # if(det(R) < 0) {
  #   Ms <- svd(M)
  #   Ms$u[,ncol(Ms$u)] <- -1 * Ms$u[,ncol(Ms$u)]
  #   M <- Ms$u %*% diag(Ms$d) %*% t(Ms$v)
  # }
  # M
}

# random nonsingular linear transformation in n dimensions
nonsingular <- function(n) matrix(rnorm(n^2), n, n)
# nonsingular.small <- function(n, e=0.01) matrix(rnorm(n^2, 0, e), n, n)
nonsingular.small <- function(n, e=0.01) diag(1, n) + matrix(rnorm(n^2, 0, e), n, n)

# cross-cosine matrix between pairs of columns of A, B
cross.cosine <- function(A, B) sapply(1:ncol(A), function(i) sapply(1:ncol(B), function(j) coop::cosine(A[,i], B[,j])))

# mcs, qmcs, smcs, imcs
mcs <- function(A, B) mean(cross.cosine(A, B))
qmcs <- function(A, B) sqrt(mean(cross.cosine(A, B)^2))
smcs <- function(A, B) {
  k <- ncol(A)
  ki <- 1/k
  cAB <- cross.cosine(A, B)
  phiA <- 1/det(coop::cosine(A))^ki
  phiB <- 1/det(coop::cosine(B))^ki
  Trace(t(cAB) %*% diag(phiB, k) %*% cAB %*% diag(phiA, k)) * ki
}
imcs <- function(A, B) {
  k <- ncol(A)
  ki <- 1/k
  cAB <- cross.cosine(A, B)
  cAAs <- svd(coop::cosine(A))
  cBBs <- svd(coop::cosine(B))
  cAA <- cAAs$v %*% t(cAAs$u)
  cBB <- cBBs$v %*% t(cBBs$u)
  Trace(t(cAB) %*% cBB %*% cAB %*% cAA) * ki
}

# anisotropy measure (geodesic distance to nearest isotropic metric)
# for example when A contains only two vectors and we measure anisotropy wrt. Theta, Iso(A) = sqrt(2) * cos(a1, a2)
# can use identity as the reference point, but not recommended -- see Moakher and Batchelor 2005
# note that the Higham08 algorithm for the matrix logarithm has trouble when M is close to identity
# we prefer to diagonalize directly if we can, and resort to Higham08 if the eigendecomposition fails
anisotropy <- function(A, scale=F, cosine=T, sqrt=F) {
  # induced metric for A
  M <- NA
  if(cosine) {
    M <- coop::cosine(A)
  } else {
    M <- crossprod(A) 
  }
  k <- ncol(A)
  
  # for the non-squared quantity...
  if(sqrt) M <- sqrtm(M)
  
  # parameter for nearest isotropic matrix
  # if using the cosine matrix, this is the distance to the equivalence class of the identity matrix
  phi <- det(M)^(1/k)
  
  # geodesic distance to isotropy
  # S <- diag(phi, k)
  # Sx <- solve(sqrtm(S))  # inefficient!
  Sx <- diag(1/sqrt(phi), k)
  aix <- tryCatch(norm(logm(Sx %*% M %*% Sx, method="Eigen"), "F"),
                  error=function(e) norm(logm(Sx %*% M %*% Sx, method="Higham08"), "F"))
  if(scale) return(aix/(1+aix))
  aix
}

# geodesic distance on PSD manifold
psd_geodesic <- function(A, B, scale=F) {
  # induced metrics
  # Sa <- solve(sqrtm(crossprod(A)))
  # B <- crossprod(B)
  Sa <- solve(sqrtm(coop::cosine(A)))
  B <- coop::cosine(B)
  
  # geodesic distance
  gd <- tryCatch(norm(logm(Sa %*% B %*% Sa, method="Eigen"), "F"),
                 error=function(e) norm(logm(Sa %*% B %*% Sa, method="Higham08"), "F"))
  if(scale) return(gd/(1+gd))
  gd
}

# scalar parameter for nearest isotropic matrix to cos(A)
psd.phi <- function(A) {
  det(coop::cosine(A))^(1/ncol(A))
}

# SPD(n) -> Corr(n)
corr_proj <- function(M) {
  dx <- 1/sqrt(diag(M))
  diag(dx) %*% M %*% diag(dx)
}

# geodesic interpolation on Corr(n) (the space of cosine similarity matrices)
# David and Gu (2019) provide the projective Riemannian geometry
# general idea is to treat Corr(n) as the quotient of SPD(n) by Diag+(n) (the space of positive diagonal matrices)
# so we just need to obtain the geodesic in SPD(n) and project to Corr(n) by the mapping X-> Diag(X)^-1/2 X Diag(X)^-1/2
# for example, the nearest isotropic matrix projects onto the identity matrix
# by default we find the geometric mean (midpoint of geodesic; see Bhatia p. 206)
corr_geodesic_interp <- function(A, B=NA, t=0.5) {
  k <- ncol(A)
  # if we don't get B, we want to interpolate relative to the identity matrix
  if(is.na(B)) {
    B <- diag(1/sqrt(psd.phi(A)), k)
  }
  psdg <- sqrtm(A) %*% expm(t * logm(solve(sqrtm(A)) %*% B %*% solve(sqrtm(A)))) %*% sqrtm(A)
  corr_proj(psdg)
}

# interpolate cross-cosine matrix toward the matrix of ones
# this isn't geometric per se
ab_interp <- function(cAB, t=0.5) {
  kr <- nrow(cAB)
  kc <- ncol(cAB)
  cAB * (1-t) + ones(kr, kc) * (t)
}

# or interpolate it toward the matrix of equal weights
# in general this gives similar results to the above method
ab_interp2 <- function(cAB, t=0.5) {
  kr <- nrow(cAB)
  kc <- ncol(cAB)
  cAB * (1-t) + matrix(rep(1/sqrt(kr*kc),kr*kc), kr, kc) * (t)
}

# compute quadratic MCS with interpolated metrics
# extra 1/geomean(ka, kb)^tab factor is because the mean denominator changes
#  - in GCD we have k squared canonical correlations
#  - in MCS we have k^2 cosine similarities
# we can also use this to see how close to isotropy we can get before cosAB is impossible
gmcs_interp <- function(A, B, ta=0.5, tb=0.5, tab=0, is.psd=F) {
  ka <- ncol(A)
  kb <- ncol(B)
  
  # compute interpolations + cross-cosine matrix
  t.aa <- corr_geodesic_interp(coop::cosine(A), t=ta)
  t.bb <- corr_geodesic_interp(coop::cosine(B), t=tb)
  t.ab <- cross.cosine(A, B)
  t.abwt <- ab_interp(t.ab, t=tab)

  # compute GCD and test PSD if needed
  gcd.interp <- 1/(sqrt(ka * kb)^tab) * (1/sqrt(ka * kb)) * Trace(t(t.abwt) %*% solve(t.bb) %*% t.ab %*% solve(t.aa))
  psd.crit <- T
  if(is.psd) {
    t.xx <- cbind(rbind(t.aa, t.ab), rbind(t(t.ab), t.bb))  # cross.cosine(A, B) makes A the rows and B the columns.
    psd.crit <- matrixcalc::is.positive.semi.definite(round(t.xx, 8))  # we check PSD here instead of PD because some of the keyword lists are nondisjoint.
  }
  return(c(gcd.interp, psd.crit))
  
  # alternative interpolation scheme for cosAB "weights"
  # t.abwt <- ab_interp2(t.ab, t=tab)
  # (1/sqrt(ka * kb)) * Trace(t(t.abwt) %*% solve(t.bb) %*% t.ab %*% solve(t.aa))
}

# function to plot GCD-MCS interpolation
plot_gmcs_interp <- function(A, B) {
  ab.interp <- t(sapply(seq(0, 1, by=0.02), function(tx) gmcs_interp(A, B, ta=tx, tb=tx, tab=tx, is.psd=T)))
  data.frame(x=seq(0, 1, by=0.02),
             y=ab.interp[,1],
             cx=ab.interp[,2]) %>%
  ggplot(aes(x, y, color=as.logical(cx))) + 
    geom_point() + 
    geom_hline(yintercept=gcd(A, B), linetype="dashed") + 
    geom_hline(yintercept=mcs(A, B), linetype="dotted") +
    scale_color_manual(values=c("tomato", "chartreuse3")) +
    theme(legend.position="none")
}

# break down by keyword list
plot_gmcs_interp2 <- function(A, B) {
  reshape2::melt(sapply(seq(0, 1, by=0.02), function(ta) sapply(seq(0, 1, by=0.02), function(tb) gmcs_interp(A, B, ta=ta, tb=tb, tab=0.25, is.psd=T)))) %>%
    ggplot(aes(x=Var1, y=Var2, fill=value)) +
    geom_tile()
}

# get summary of interpolation
# 0.02 is reasonably fast + precise; could always set the granularity of this to be finer
summ_gmcs_interp <- function(A, B, An=NA, Bn=NA) {
  ab.interp <- t(sapply(seq(0, 1, by=0.02), function(tx) gmcs_interp(A, B, ta=tx, tb=tx, tab=tx, is.psd=T)))
  data.frame(x=seq(0, 1, by=0.02),
             y=ab.interp[,1],
             cx=ab.interp[,2],
             An=An,
             Bn=Bn)
}

# how far apart are the orientation components?
# this gives the sin of the angle of rotation between them;
#  in SO(3) this is sin(acos((Tr(A'B) - 1)/2)) e.g.
# this is a factor of sqrt(2) off from the projection distance ||Pa-Pb||_2
# NOTE: technically this is for SO(n) so they have to be det = +1
orientation.angle <- function(A, B) {
  sqrt(sum(unique(abs(Im(eigen(t(A) %*% B)$values)))^2))
}

# the trace product Tr(Ua'Ub) is also interesting (the between angle) since we can also view this
# as the log conditional probability p(B|A) if A, B are spherical; the other component determines
# the cosine of the central angle subtended by the span of A: 0F1(m/2, S^2/4)
```


```{r load_data, echo=F, message=F, warning=F}
# load GloVe embeddings
glove.dC <- read_delim("/Users/akindel/Code/common/GloVe/wikipedia_0923/C_vectors.txt", " ", col_names = F)
glove.dW <- read_delim("/Users/akindel/Code/common/GloVe/wikipedia_0923/W_vectors.txt", " ", col_names = F)
glove.wf <- read_delim("/Users/akindel/Code/common/GloVe/wikipedia_0923/vocab.txt", " ", col_names = c("word", "frequency"))

# clean up
words.g <- glove.dC$X1
bias_C <- glove.dC$X302
bias_W <- glove.dW$X302
mcx <- mc <- as.matrix(glove.dC[-c(1,302)])
mwx <- mw <- as.matrix(glove.dW[-c(1,302)])
rm(glove.dC, glove.dW)

# optionally: orthogonalize and ignore top 3 singular vectors
# sometimes there is a dominant "frequency dimension"; this reduces that phenomenon to some degree
# whether this is a good thing depends on the application
# for example in 2010s-era "word analogy tasks" the test words all have similar marginal frequencies
# mcsvd <- svd(mc)
# mwsvd <- svd(mw)
# mc <- mcsvd$u[,4:300]
# mw <- mwsvd$u[,4:300]

# another method is to project it onto the orthogonal complement of the top 3 singular vectors
# not 100% sure what this adds over dropping 
# mcsvd <- svd(mc)
# mwsvd <- svd(mw)
# mc <- mc %*% (diag(1, 300) - tcrossprod(mcsvd$v[,1:3]))
# mw <- mw %*% (diag(1, 300) - tcrossprod(mwsvd$v[,1:3]))

# a bigger problem with this representation is that the dimensions are not identified by the model
# to address this we can estimate the varimax rotation to put the embeddings into a "sparse orientation"
# see Rohe and Zeng 2023 JRSSB for the corresponding semiparametric latent factor model
# note that we don't apply the "Kaiser normalization" per Rohe and Zeng
# to speed this up it usually works really well to use only the most common words to estimate the rotation
# here I'm using 40k which is about 10% of the vocabulary
# for example in the Wikipedia data this is approximately 95% of the word occurrences 
# note that you then need the cross-encoding matrix RwtRc to preserve reconstruction error
# per Rohe and Zeng this is sort of like the factor "intercorrelations"
# this doesn't actually change anything in this paper but you should do it :)
vn.est <- 40000
mc.vr <- varimax(mc[1:vn.est,], F)$rotmat
mw.vr <- varimax(mw[1:vn.est,], F)$rotmat
mwc.B <- t(mw.vr) %*% mc.vr

# apply varimax orientations
mcv <- mc %*% mc.vr
mwv <- mw %*% mw.vr

# represent words as column vectors
# this makes the connection to CCA a little easier to see
mc <- t(mc)
mw <- t(mw)
mcv <- t(mcv)
mwv <- t(mwv)

# remember that SVD is invariant to paired sign permutations of the columns of U, V
# this is true regardless of whether we estimate an identified rotation matrix
# we can orient embedding dimensions with consistent signs by making all the heavy tails positive
# we then push the reorienting permutations through to the center B matrix
mcv.dir <- sapply(1:nrow(mcv), function(j) sign(mcv[j,which.max(abs(mcv[j,]))]))
mwv.dir <- sapply(1:nrow(mwv), function(j) sign(mwv[j,which.max(abs(mwv[j,]))]))
mcv <- diag(mcv.dir) %*% mcv
mwv <- diag(mwv.dir) %*% mwv
mwc.B <- diag(mwv.dir) %*% mwc.B %*% diag(mcv.dir)

# construct dimension labels (ten words with largest "loadings")
colnames(mcv) <- colnames(mwv) <- colnames(mc) <- colnames(mw) <- names(bias_C) <- names(bias_W) <- words.g
mcv.dimlabels <- sapply(1:nrow(mcv), function(j) paste0(names(sort(mcv[j,], T)[1:10]), collapse=", "))
mwv.dimlabels <- sapply(1:nrow(mwv), function(j) paste0(names(sort(mwv[j,], T)[1:10]), collapse=", "))

# sum of "loadings" for top 10 labels (rough diagnostic for how identified it is?)
mcv.labelscores <- sapply(1:nrow(mcv), function(j) sum(sort(mcv[j,], T)[1:10]))
mwv.labelscores <- sapply(1:nrow(mwv), function(j) sum(sort(mwv[j,], T)[1:10]))

# view intercorrelations of factors
# the sign of these is hard to interpret IMO; maybe square it?
data.frame(reshape2::melt(mwc.B)) %>%
  left_join(data.frame(Var1 = 1:nrow(mwv), mwv.dimlabels)) %>%
  left_join(data.frame(Var2 = 1:nrow(mcv), mcv.dimlabels)) %>%
  left_join(data.frame(Var1 = 1:nrow(mwv), mwv.labelscores)) %>%
  left_join(data.frame(Var2 = 1:nrow(mcv), mcv.labelscores)) %>%
  mutate(value2 = value^2) ->
  mwc.Bcorr
```


```{r keywords, eval=F}
# filter to sample vocabulary (uniquely)
vocabulary.filter <- function(wl) words.g[words.g %in% unlist(c(wl))]

# keyword lists (WEAT)
flowers <- c(str_split("aster, clover, hyacinth, marigold, poppy, azalea, crocus, iris, orchid, rose, bluebell, daffodil, lilac, pansy, tulip, buttercup, daisy, lily, peony, violet, carnation, gladiolus, magnolia, petunia, zinnia", ", ", Inf, T))  # "gladiola" -> "gladiolus"
insects <- c(str_split("ant, caterpillar, flea, locust, spider, bedbug, centipede, fly, maggot, tarantula, bee, cockroach, gnat, mosquito, termite, beetle, cricket, hornet, moth, wasp, blackfly, dragonfly, horsefly, roach, weevil", ", ", Inf, T)) 
pleasant <- c(str_split("caress, freedom, health, love, peace, cheer, friend, heaven, loyal, pleasure, diamond, gentle, honest, lucky, rainbow, diploma, gift, honor, miracle, sunrise, family, happy, laughter, paradise, vacation", ", ", Inf, T))
unpleasant <- c(str_split("abuse, crash, filth, murder, sickness, accident, death, grief, poison, stink, assault, disaster, hatred, pollute, tragedy, divorce, jail, poverty, ugly, cancer, kill, rotten, vomit, agony, prison", ", ", Inf, T))
# pleasant_b <- c(str_split("joy, love, peace, wonderful, pleasure, friend, laughter, happy", ", ", Inf, T))
# unpleasant_b <- c(str_split("agony, terrible, horrible, nasty, evil, war, awful, failure", ", ", Inf, T))
instruments <- c(str_split("bagpipe, cello, guitar, lute, trombone, banjo, clarinet, harmonica, mandolin, trumpet, bassoon, drum, harp, oboe, tuba, bell, fiddle, harpsichord, piano, viola, bongo, flute, horn, saxophone, violin", ", ", Inf, T))
weapons <- c(str_split("arrow, club, gun, missile, spear, ax, dagger, harpoon, pistol, sword, blade, dynamite, hatchet, rifle, tank, bomb, firearm, knife, shotgun, teargas, cannon, grenade, mace, slingshot, whip", ", ", Inf, T))  # "axe" -> "ax"
career <- c(str_split("executive, management, professional, corporation, salary, office, business, career", ", ", Inf, T))
family <- c(str_split("home, parents, children, family, cousins, marriage, wedding, relatives", ", ", Inf, T))
temporary <- c(str_split("impermanent, unstable, variable, fleeting, short, brief, occasional", ", ", Inf, T))  # short-term --> short
permanent <- c(str_split("stable, always, constant, persistent, chronic, prolonged, forever", ", ", Inf, T))
math <- c(str_split("math, algebra, geometry, calculus, equations, computation, numbers, addition", ", ", Inf, T))
arts <- c(str_split("poetry, art, dance, literature, novel, symphony, drama, sculpture, shakespeare", ", ", Inf, T))  # combined with second arts list (adds "shakespeare"))
science <- c(str_split("science, technology, physics, chemistry, einstein, nasa, experiment, astronomy", ", ", Inf, T))  # lowercased "Einstein" and "NASA"
male <- c(str_split("brother, father, uncle, grandfather, son, he, his, him", ", ", Inf, T))
female <- c(str_split("sister, mother, aunt, grandmother, daughter, she, hers, her", ", ", Inf, T))
mental_illness <- c(str_split("sad, hopeless, gloomy, tearful, miserable, depressed", ", ", Inf, T))
physical_illness <- c(str_split("sick, illness, influenza, disease, virus, cancer", ", ", Inf, T))
male_names <- c(str_split(tolower("John, Paul, Mike, Kevin, Steve, Greg, Jeff, Bill"), ", ", Inf, T))
female_names <- c(str_split(tolower("Amy, Joan, Lisa, Sarah, Diana, Kate, Ann, Donna"), ", ", Inf, T))
white_names <- c(str_split(tolower("Brad, Brendan, Geoffrey, Greg, Brett, Jay, Matthew, Neil, Todd, Allison, Anne, Carrie, Emily, Jill, Meredith, Sarah, Kristen, Laurie"), ", ", Inf, T))
black_names <- c(str_split(tolower("Darnell, Hakim, Jermaine, Kareem, Jamal, Leroy, Rasheed, Tremayne, Tyrone, Aisha, Ebony, Keisha, Kenya, Latoya, Latanya, Tamika, Tanisha"), ", ", Inf, T))  # removed "Lakisha"
young_names <- c(str_split(tolower("Tiffany, Michelle, Cindy, Kristy, Brad, Eric, Joey, Billy"), ", ", Inf, T))  # The female names here seem pretty outdated to me.
old_names <- c(str_split(tolower("Ethel, Bernice, Gertrude, Agnes, Cecil, Wilbert, Mortimer, Edgar"), ", ", Inf, T))

# keyword lists (KTE 2019 ASR)
affluence <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/affluence_pairs.csv", col_names = c("RICH", "POOR"))
gender <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/gender_pairs.csv", col_names = c("MASCULINE", "FEMININE"))
cultivation <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/cultivation.csv")
education <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/education.csv")
employment <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/employment.csv")
morality <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/morality.csv")
status <- read_csv("/Users/akindel/Code/cosine/data/kte2019/word_pairs/status.csv")
rich <- vocabulary.filter(affluence[,1])
poor <- vocabulary.filter(affluence[,2])
masculine <- vocabulary.filter(gender[,1])
feminine <- vocabulary.filter(gender[,2])
civilized <- vocabulary.filter(cultivation[,1])
uncivilized <- vocabulary.filter(cultivation[,2])
educated <- vocabulary.filter(education[,1])
uneducated <- vocabulary.filter(education[,2])
boss <- vocabulary.filter(employment[,1])
worker <- vocabulary.filter(employment[,2])
good <- vocabulary.filter(morality[,1])
evil <- vocabulary.filter(morality[,2])
status_high <- vocabulary.filter(status[,1])
status_low <- vocabulary.filter(status[,2])

# list of keyword lists
# really not every pairwise comparison of these is theoretically interesting!
# however, it's indicative of the structure of the task people are doing
# the main problem with some of these is that they're quite short...
wordlists <- tibble::lst(flowers, insects, pleasant, unpleasant, instruments, weapons, career, family, temporary, permanent, math, arts, science, male, female, mental_illness, physical_illness, male_names, female_names, white_names, black_names, young_names, old_names, rich, poor, masculine, feminine, civilized, uncivilized, educated, uneducated, boss, worker, good, evil, status_high, status_low)
wl.labels <- names(wordlists)
```


```{r keywords_anisotropy, eval=F}
# evaluate anisotropy for k = length(wl)
# choosing words uniformly at random from the sample vocabulary
# every keyword list is much more anisotropic than random keyword lists
# also, random keyword lists are much more anisotropic than random vectors
nsp <- 2000
sapply(wordlists, function(x) anisotropy(mcv[,x])) -> aix.test.ref
sapply(wordlists, function(x) replicate(nsp, anisotropy(mcv[,sample(1:ncol(mcv), length(x))])) ) -> aix.test1

# plot observed anisotropy against expected anisotropy of a random keyword list
# interval shows 99% of samples under null distribution
data.frame(rn=1:length(wordlists),
           aix.exp=colMeans(aix.test1),
           aix.obs=aix.test.ref,
           wl.k=sapply(wordlists, function(w) length(w)),
           label=paste0(names(wordlists), " (", sapply(wordlists, function(w) length(w)), ")")) %>%
  mutate(wl.r = rank(wl.k, ties.method="first")) %>%
  mutate(label.r = factor(wl.r, levels=wl.r, labels=label)) %>%
  rowwise() %>%
  mutate(aix.xl = quantile(aix.test1[,rn], 0.005),
         aix.xu = quantile(aix.test1[,rn], 0.995)) %>%
  arrange(wl.r) -> 
  aix.test.df

aix.test.df %>%
  ggplot(aes(y=wl.r)) + 
  geom_point(aes(x=aix.exp)) + 
  geom_point(aes(x=aix.obs), color="tomato", shape=19, size=3) + 
  geom_errorbarh(aes(xmin=aix.xl, xmax=aix.xu)) +
  scale_y_continuous(labels=str_replace_all(aix.test.df$label.r, "_", " "), breaks=1:37) + xlim(0, NA) +
  labs(x="Isotropy index",
       y="Keyword list (# words)") -> 
  p1

p1
```


```{r perturb_list_direct, eval=F}
# we can also directly perturb the positions of one or more keywords
#  *numerically* by applying a small rotation to the chosen positions
# by default we just rotate A but we can do both simultaneously too.
mcs_perturb <- function(A, B, Ae, Be=0, ns=F) {
  # generate rotation
  Ma <- NA
  Mb <- NA
  if(ns) {
    Ma <- nonsingular.small(ncol(A), e=Ae)
    Mb <- nonsingular.small(ncol(B), e=Be)
  } else {
    Ma <- rotator.small(ncol(A), e=Ae) * -1
    Mb <- rotator.small(ncol(B), e=Be) * -1
  }
  AM <- A %*% Ma
  BM <- B %*% Mb
  
  # how big is this perturbation?
  # mag <- diag(cross.cosine(AM, A))^2
  # mean.mag <- mean(mag)
  # max.mag <- max(mag)
  # min.mag <- min(mag)
  
  # mcs and gcd
  # Acs <- mean(coop::cosine(A)[upper.tri(diag(1, ncol(A)))])
  # AMcsM <- coop::cosine(AM)
  # AMcs <- mean(AMcsM[upper.tri(diag(1, ncol(A)))])
  # csx <- cross.cosine(AM, B)
  mcsx <- mcs(AM, BM)
  gcdx <- gcd(AM, BM)  # for completeness
  
  # anisotropy diagnostic
  aix.am <- anisotropy(AM)
  aix.bm <- anisotropy(BM)
  
  # phi diagnostic
  # phi.am <- psd.phi(AM)
  # phi.bm <- psd.phi(BM)
  
  # unused diagnostics
  # kmo.A <- psych::KMO(coop::cosine(AM))
  # kmo.B <- psych::KMO(coop::cosine(BM))
  
  # return(c(mcs=mcsx, gcd=gcdx, mean.ang=mean.mag, max.ang=max.mag, min.ang=min.mag, aix.A=aix.am, aix.B=aix.bm, phi.A=phi.am, phi.B=phi.bm, Ae=Ae, Be=Be))
  return(c(mcs=mcsx, gcd=gcdx, aix.A=aix.am, aix.B=aix.bm, Ae=Ae, Be=Be))
}

# simulation for one pair of keyword lists
ldperturb <- function(Ai, Bi, nsamp=100) {
  A <- wordlists[[Ai]]
  B <- wordlists[[Bi]]
  test.baseline <- mcs(mcv[,A], mcv[,B])
  aix.baseline <- anisotropy(mcv[,A]) + anisotropy(mcv[,B])
  test.nsx <- lapply(seq(0.01, 1, by=0.05), function(ex) data.frame(t(replicate(nsamp, mcs_perturb(mcv[,A], mcv[,B], Ae=ex, Be=ex, ns=T))), 
                                                                      Al=names(wordlists)[Ai], Bl=names(wordlists)[Bi], tbx=test.baseline, axx=aix.baseline) )
  test.ns.df <- do.call(rbind.data.frame, test.nsx) %>% mutate(aix.s = aix.A + aix.B) %>% group_by(Ae) %>% mutate(mcs.m = mean(mcs), aix.m = mean(aix.s)) %>% ungroup()
  return(test.ns.df)
}

# function to plot results from one simulation (one list pair)
plot.ldperturb <- function(tdf) {
  tdf %>%
  ggplot(aes(x=aix.A + aix.B, y=mcs, color=Ae)) +
  geom_point() +
  geom_hline(yintercept=unique(tdf$tbx), linetype="dashed") +
  geom_vline(xintercept=unique(tdf$axx), linetype="dotted") +
  scale_color_viridis_c() +
  labs(x="Iso(A) + Iso(B)",
       y="MCS(A, B)",
       color="Size of transformation") +
  theme(legend.position="bottom",
        legend.text = element_text(size=12),
        legend.key.size=unit(0.8, "cm"))
}

# run perturbation simulation over all possible keyword list pairs
# this takes about 36 hours on my machine
# kwl.perturb <- sapply(1:(length(wordlists)-1), function(w1i) sapply(w1i:length(wordlists), function(w2i) ldperturb(w1i, w2i) %>% distinct(Ae, Be, Al, Bl, mcs.m, aix.m) ))
# do.call(rbind.data.frame, lapply(1:length(kwl.perturb), function(i) {
#     L <- kwl.perturb[[i]]
#     Lx <- lapply(1:(length(L)/6), function(j) {
#         M <- do.call(cbind.data.frame, kwl.perturb[[i]][((j-1)*6 + 1):(j*6)])
#         colnames(M) <- c("Ae", "Be", "Al", "Bl", "mcs.m", "aix.m")
#         M
#     } )
#     do.call(rbind.data.frame, Lx)
# })) %>% filter(Al != Bl) -> kwl.perturb.df

# plot 
kwl.perturb.df %>%
  ggplot(aes(x=aix.m, y=mcs.m, color=Ae)) + 
  geom_point() + 
  scale_color_viridis_c() +
  labs(x="Mean anisotropy",
       y="Mean MCS",
       color="Size of transformation") +
  theme(legend.position="bottom",
        legend.text = element_text(size=12),
        legend.key.size=unit(0.8, "cm")) ->
  p2

p2

# plot some example associations
p3a <- plot.ldperturb(ldperturb(31, 25, nsamp=100)) + ggtitle("uneducated X poor")
p3b <- plot.ldperturb(ldperturb(26, 4, nsamp=100)) + ggtitle("masculine X unpleasant") 
# gridExtra::grid.arrange(p3a, p3b, nrow=1)
ggpubr::ggarrange(p3a, p3b, nrow=1, legend="bottom", common.legend = T)
```


```{r mcs_gcd_interpolation, eval=F}
# summarize all interpolations
# we're mainly interested in the SPD breakdown point
# this tells us how far MCS is from geometric consistency (1 means MCS is consistent with the geometry of A and B)
# anything less than 1 means MCS is impossible, but higher values (brighter colors) are closer to consistency
# this code is pretty inefficient :(
kwl.interp <- lapply(1:length(wordlists), function(w1i) lapply(1:length(wordlists), function(w2i) summ_gmcs_interp(mcv[,wordlists[[w1i]]], mcv[,wordlists[[w2i]]], wl.labels[w1i], wl.labels[w2i]) ))
do.call(rbind, do.call(rbind, kwl.interp)) %>% 
  distinct_all() %>% 
  filter(as.logical(cx)) %>% 
  group_by(An, Bn) %>% 
  summarize(max.x = max(x)) %>% 
  ungroup() %>% 
  complete(An, Bn, fill=list(max.x=NA)) -> # %>%
  # filter(An != Bn) ->
  kwl.interp.df

# proportion of keyword list pairs for which cosAB is compatible with exactly isotropic cosAA, cosBB
sum(kwl.interp.df$max.x == 1)/nrow(kwl.interp.df)

# proportion of keyword list pairs for which the interpolation is nonmonotonic
# i.e. for how many comparisons is the minimum value of y not at t = 0 or 1?
do.call(rbind, do.call(rbind, kwl.interp)) %>% distinct_all() %>% filter(An != Bn) %>% group_by(An, Bn) %>% filter(y == min(y)) %>% distinct(y, .keep_all=T) %>% mutate(mono.x = x %in% c(0, 1)) %$% (1 - mean(mono.x))

# plot grid of SPD breakdown points
# kwl.interp.df %>%
#   ggplot(aes(x=str_replace_all(An, "_", " "), y=str_replace_all(Bn, "_", " "), fill=max.x)) +
#   geom_tile() +
#   scale_fill_viridis_c(limits=c(0, 1)) +
#   labs(x="List A",
#        y="List B",
#        fill="Breakdown point") +
#   theme(legend.position="bottom",
#         axis.text.x = element_text(angle=45, hjust=1)) ->
#   p4

# assign labels in order of cardinality (clunky...)
kwl.interp.df %>% 
  rowwise() %>% mutate(An.k = length(wordlists[[An]]), Bn.k = length(wordlists[[Bn]])) %>% ungroup() %>% 
  group_by(An) %>% mutate(Bn.r = rank(Bn.k, ties.method="first")) %>% ungroup() %>% 
  group_by(Bn) %>% mutate(An.r = rank(An.k, ties.method="first")) %>% ungroup() %>% 
  mutate(label.An = factor(An.r, levels=An.r, labels=An), label.Bn = factor(Bn.r, levels=Bn.r, labels=Bn)) %>% ungroup() %>% 
  arrange(An.r, Bn.r) %>% 
  mutate(max.x = ifelse(An == Bn, NA, max.x)) -> 
  kwl.interp.df2

kwl.interp.df2 %>%
  ggplot(aes(x=An.r, y=Bn.r, fill=max.x)) + 
  geom_tile() +
  scale_fill_viridis_c(limits=c(0, 1), na.value = NA) +
  scale_x_continuous(labels=str_replace_all(unique(kwl.interp.df2$An), "_", " "), breaks=1:37, expand=c(0, 0)) +
  scale_y_continuous(labels=str_replace_all(unique(kwl.interp.df2$Bn), "_", " "), breaks=1:37, expand=c(0, 0)) +
  labs(x="List A",
       y="List B",
       fill="Breakdown point") +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle=45, hjust=1),
        legend.text = element_text(size=12),
        legend.key.size=unit(0.8, "cm")) ->
  p4
p4

# plot some "typical" example interpolations between MCS and GCD
# x=0 is GCD, x=1 is MCS
plot_gmcs_interp(mcv[,wordlists[[24]]], mcv[,wordlists[[35]]]) + labs(x="t", y="Association measure", title=paste0(wl.labels[24], " X ", wl.labels[35])) -> p5a
plot_gmcs_interp(mcv[,wordlists[[31]]], mcv[,wordlists[[37]]]) + labs(x="t", y="Association measure", title=paste0(wl.labels[31], " X ", "status low")) -> p5b
# gridExtra::grid.arrange(p5a, p5b, nrow=1)
ggpubr::ggarrange(p5a, p5b, nrow=1, legend="none")

# some monotonic ones
# plot_gmcs_interp(mcv[,wordlists[[17]]], mcv[,wordlists[[13]]]) + labs(x="t", y="Association measure", title=paste0(wl.labels[17], " X ", wl.labels[13])) -> p5.3
# plot_gmcs_interp(mcv[,wordlists[[31]]], mcv[,wordlists[[13]]]) + labs(x="t", y="Association measure", title=paste0(wl.labels[31], " X ", wl.labels[13])) -> p5.4
# gridExtra::grid.arrange(p5.1, p5.2, p5.3, p5.4, nrow=2)
```



\pagebreak

# Appendix {-}

## A. Keyword lists {-}

Keyword lists are drawn from @caliskan2017semantics and @kozlowski2019geometry. A small number of modifications to the lists are needed to ensure coverage in the sample vocabulary; see rightmost column.

```{r, eval=F, warning=F, message=F, echo=F}
# main script defines a variable "wordlists" with all the keyword lists
# print this out programmatically if needed
sapply(wordlists, function(wl) paste0(wl, collapse=", "))
```

\singlespacing
\renewcommand*{\arraystretch}{1.4}
\begin{center}
\begin{longtable}{|p{0.2\linewidth}|p{0.55\linewidth}|p{0.25\linewidth}|}
 Label & Keywords & Modifications \\ \hline
 flowers & aster, clover, hyacinth, marigold, poppy, azalea, crocus, iris, orchid, rose, bluebell, daffodil, lilac, pansy, tulip, buttercup, daisy, lily, peony, violet, carnation, gladiolus, magnolia, petunia, zinnia & "gladiola" $\rightarrow$ "gladiolus" (spelling) \\ 
 insects & ant, caterpillar, flea, locust, spider, bedbug, centipede, fly, maggot, tarantula, bee, cockroach, gnat, mosquito, termite, beetle, cricket, hornet, moth, wasp, blackfly, dragonfly, horsefly, roach, weevil & \\ 
 pleasant & caress, freedom, health, love, peace, cheer, friend, heaven, loyal, pleasure, diamond, gentle, honest, lucky, rainbow, diploma, gift, honor, miracle, sunrise, family, happy, laughter, paradise, vacation & \\ 
 unpleasant & abuse, crash, filth, murder, sickness, accident, death, grief, poison, stink, assault, disaster, hatred, pollute, tragedy, divorce, jail, poverty, ugly, cancer, kill, rotten, vomit, agony, prison & \\ 
 instruments & bagpipe, cello, guitar, lute, trombone, banjo, clarinet, harmonica, mandolin, trumpet, bassoon, drum, harp, oboe, tuba, bell, fiddle, harpsichord, piano, viola, bongo, flute, horn, saxophone, violin & \\ 
 weapons & arrow, club, gun, missile, spear, ax, dagger, harpoon, pistol, sword, blade, dynamite, hatchet, rifle, tank, bomb, firearm, knife, shotgun, teargas, cannon, grenade, mace, slingshot, whip & "axe" $\rightarrow$ "ax" (spelling) \\ 
 career & executive, management, professional, corporation, salary, office, business, career & \\ 
 family & home, parents, children, family, cousins, marriage, wedding, relatives & \\ 
 temporary & impermanent, unstable, variable, fleeting, short, brief, occasional & "short-term" $\rightarrow$ "short" (pre-processing compatibility) \\ 
 permanent & stable, always, constant, persistent, chronic, prolonged, forever & \\ 
 math & math, algebra, geometry, calculus, equations, computation, numbers, addition & \\ 
 arts & poetry, art, dance, literature, novel, symphony, drama, sculpture, shakespeare & Added "shakespeare" (merged two high-overlap "arts" lists) \\ 
 science & science, technology, physics, chemistry, einstein, nasa, experiment, astronomy & \\ 
 male & brother, father, uncle, grandfather, son, he, his, him & \\ 
 female & sister, mother, aunt, grandmother, daughter, she, hers, her & \\
 mental illness & sad, hopeless, gloomy, tearful, miserable, depressed & \\ 
 physical illness & sick, illness, influenza, disease, virus, cancer & \\ 
 male names & john, paul, mike, kevin, steve, greg, jeff, bill & \\ 
 female names & amy, joan, lisa, sarah, diana, kate, ann, donna & \\ 
 white names & brad, brendan, geoffrey, greg, brett, jay, matthew, neil, todd, allison, anne, carrie, emily, jill, meredith, sarah, kristen, laurie & \\ 
 black names & darnell, hakim, jermaine, kareem, jamal, leroy, rasheed, tremayne, tyrone, aisha, ebony, keisha, kenya, latoya, latanya, tamika, tanisha & Dropped "lakisha" (below frequency threshold in Wikipedia corpus) \\ 
 young names & tiffany, michelle, cindy, kristy, brad, eric, joey, billy & \\ 
 old names & ethel, bernice, gertrude, agnes, cecil, wilbert, mortimer, edgar & \\ 
 rich & developed, successful, rich, wealthy, valuable, expensive, luxury, precious, costly, prosperous, richest, affluent, privileged, solvent, lavish, luxurious, flush, richer, extravagant, upscale, exquisite, invaluable, sumptuous, plush, priceless, opulent, posh, classy, exorbitant, affluence, opulence, solvency, luxuriant, advantaged, moneyed, propertied, swanky, ritzy & \\ 
 poor & poor, basic, poverty, plain, unsuccessful, cheap, ruined, bankrupt, economical, impoverished, inexpensive, disadvantaged, poorer, poorest, needy, insolvency, destitute, worthless, underprivileged, underdeveloped, insolvent, penniless, indigent, squalid, ramshackle, threadbare, penury, valueless, impecunious, skint, indigence, penurious, moneyless, necessitous, beggarly & \\ 
 masculine & he, his, him, man, men, male, boys, boy, masculine & \\ 
 feminine & she, her, women, female, woman, girls, girl, feminine, hers & \\ 
 civilized & civil, formal, proper, cultivated, refined, polished, cultured, polite, civilized, cordial, gracious, courtly, mannered, courteous, civility, genteel, obliging, urbane, urbanity, edified & \\ 
 uncivilized & informal, blunt, improper, rude, gruff, uncultivated, unpolished, uncivilized, unrefined, rudeness, boorish, uncultured, incivility, uncivil, discourteous, loutish, ungracious & \\ 
 educated & taught, educated, trained, learned, literate, knowledgeable, lettered, schooled, tutored & \\ 
 uneducated & ignorant, illiterate, untrained, uneducated, unlearned, unschooled, unlettered, untutored, untaught & \\ 
 boss & director, manager, owner, directors, owners, boss, superintendent, managers, employers, employer, supervisor, capitalist, industrialist, foreman, proprietor, bosses, proprietors, industrialists, capitalists, foremen & \\ 
 worker & staff, workers, employees, worker, employee, laborers, proletarian, laborer, proletariat & \\ 
 good & good, holy, moral, noble, pure, virtue, ethical, honest, decent, righteous, valiant, virtuous, dignified, chaste, altruistic, principled, commendable, scrupulous, chivalrous, unquestionable, upstanding, guiltless & \\ 
 evil & vice, bad, guilty, evil, corrupt, wicked, questionable, crooked, selfish, immoral, villainous, indecent, unethical, sinful, dishonest, unscrupulous, impure, unholy, nefarious, transgressive, reprehensible, odious, unsavory, fiendish, unprincipled, undignified & \\ 
 status high & prominent, distinguished, influential, renowned, prestigious, acclaimed, eminent, honorable, illustrious, venerable, esteemed, exalted, reputable, dignitary, estimable & \\ 
 status low & common, ordinary, modest, humble, commonplace, mundane, commoner, lowly, unpretentious, disreputable, prosaic, dishonorable & 
\end{longtable}
\end{center}

## B. Notation {-}

- $A^{-1}$, the matrix inverse of A.
- $A^{1/2}$, the matrix square root of A, i.e., the matrix B such that $BB = A$.
- $A^T$, the transpose of A.
- $\text{Diag}(A)$, the matrix with i-th diagonal entry $X_{ii} = A_{ii}$ and off-diagonal entries equal to zero.
- $\text{tr}(A)$, the trace of A.
- $\text{det}(A)$, the determinant of A.
- $\text{Exp}(A)$, the matrix exponential of A, i.e. $\sum_{k = 0}^\infty \frac{1}{k!} A^k$.
- $\text{Log}(A)$, the matrix logarithm of A.
- $||A||_F$, the Frobenius norm of A, i.e. $\sqrt{\text{tr}(A^T A)}$.
- $\left<a, b\right>$, the inner product of vectors $a$ and $b$.

## C. Software {-}

All results are computed in R (4.3.1). Code to reproduce the results requires `tidyverse` (2.0.0) and `pracma` (2.4.2) and their dependencies.

